### 2018-07-26 paper/paper.tex
First commit

\documentclass[11pt,a4paper]{article}

\usepackage[margin=1in]{geometry}

\title{Cellular Automata and Computational Universality}

\begin{document}

\maketitle

\begin{center}
    Thomas Archbold \\
    University of Warwick \\
    \texttt{T.Archbold@warwick.ac.uk}
\end{center}

\begin{abstract}
    Cellular automata are discrete models with the ability to not only give rise
    to beautiful, intricate patterns, but also to be used as powerful tools of
    computation, with applications in cryptography, error-correction coding, and
    simulation of computer processors, to name a few. They also raise profound
    questions about the nature of our reality, asking whether our universe could
    be one such automaton. This paper provides a discussion into these automata,
    exploring the various power and limitations of a number of specific rule
    sets, covering John Conway's well-known ``Game of Life'' to the more obscure
    ``Langton's ant'' and ``Wireworld''. In particular, it explores the notion
    of computational universality, or Turing completeness, an automaton's
    ability to simulate any conceivable computation, and considers their
    potential in the context of solving two specific problems, the Firing Squad
    Synchronization Problem and the Majority Problem. Existing solutions to
    these problems are explored, and their existing avenues for optimization are
    discussed. In order to fully appreciate the complex structures that can
    arise from such simple beginnings, this project also presents software to
    visualize and probe further into the nature of the automata highlighted.
\end{abstract}

\end{document}

### 2018-07-30 paper/bibliography.tex
Visualisation working

John E. Hopcroft, Rajeev Motwani, Jeffrey D. Ullman, 'Introduction to Automata
Theory, Languages, and Computation', 3rd Edition, Prentice Hall, 2006

Hiroshi Umeo, Kinuo Nishide, Keisuke Kubo, 'A Simple Optimum-Time FSSP Algorithm
for Multi-Dimensional Cellular Automata', journal ref: EPTCS 90, 2012, pp.
151-165 (arXiv:1208.2761)

J. Gruska, S. La Torre, M. Napoli, M. Parente, 'Various Solutions to the Firing
Squad Synchronization Problems', 2005 (arXiv:cs/0511044)

Danny Dolev, Ezra N. Hoch, Yoram Moses, 'An Optimal Self-Stabilizing Firing
Squad', 2009 (arXiv:0908.2295)

Nazim Fatès, Irène Marcovici, Siamak Taati, 'Two-dimensional traffic rules and
the density classification problem', 2016 (arXiv:1604.04402)

Henryk Fuks, 'Solution of the Density Classification Problem with Two Cellular
Automata Rules', journal ref: Phys. Rev. E 55, 2081R (1997)
(arXiv:comp-gas/9703001)

### 2018-07-30 paper/paper.tex
Visualisation working

@@
 \usepackage[margin=1in]{geometry}
+\usepackage{amsmath}
 
@@
 
+\section{Introduction}
+    A cellular automaton is a discrete model of computation which consists of a
+    finite collection of ``cells'', each in one of a finite number of states.
+    The state of each of these cells may evolve over the discrete progression of
+    time in discrete steps, and may do so according to a deterministic set of
+    rules specifying the state to which each cell is to progress, taking into
+    account the states of cells in its neighbourhood. Their inherently discrete
+    nature allows for strong analogies to be made with digital computers, and
+    gifts them the ability to simulate digital processes and the potential to
+    solve problems in this area.
+
+    Consider the cellular automaton defined by the simple rule:
+    $$ a_{i}^{t+1} = a_{i-1}^{t} + a_{i+1}^{t} \mod{2} $$
+    For any automaton, in each time step the rule is applied to all cells in the
+    automaton instantaneously and simultaneously. In this case, our set of
+    states is ${0,1}$, and for each cell we look at the cells immediately
+    preceding and succeeding it: if exactly one of them is in state 1, then this
+    cell will be in state 1 in the next time step. Otherwise it will be in state
+    0.
+
+\section{Examples}
+\section{Applications}
+\section{Computational Universality}
+\section{Firing Squad Synchronization Problem}
+\section{Majority Problem}
+\section{}
 \end{document}

### 2018-07-30 simulation/SDL/include/automaton.h

@@ -1,0 +1,19 @@
+#ifndef _AUTOMATON_H
+#define _AUTOMATON_H
+
+#include <SDL2/SDL.h>
+
+struct automaton;
+typedef void (*simulate_fn)(struct automaton *);
+
+struct automaton {
+    int *cells;
+    SDL_Rect *rects;
+    int len;
+    int dimension;
+    simulate_fn sim;
+};
+
+struct automaton *init_automaton(int len, simulate_fn func, int d);
+
+#endif

### 2018-07-30 simulation/SDL/include/global.h

@@ -1,0 +1,10 @@
+#ifndef _GLOBAL_H
+#define _GLOBAL_H
+
+#define WIN_WIDTH 640
+#define WIN_HEIGHT 480
+
+#define ALIVE 1
+#define DEAD 0
+
+#endif

### 2018-07-30 simulation/SDL/src/automaton.c

@@ -1,0 +1,23 @@
+#include <stdlib.h>
+#include "global.h"
+#include "automaton.h"
+
+struct automaton *init_automaton(int len, simulate_fn func, int d) {
+    struct automaton *ca = malloc(sizeof(struct automaton));
+    ca->cells = calloc(len, sizeof(int));
+    
+    ca->len = len;
+    ca->dimension = d;
+    ca->sim = func;
+
+    ca->rects = malloc(sizeof(SDL_Rect) * len);
+
+    for (int i = 0; i < len; i++) {
+        ca->rects[i].w = WIN_WIDTH / len;
+        ca->rects[i].h = ca->rects[i].w;
+        ca->rects[i].x = i * ca->rects[i].w;
+        ca->rects[i].y = 0;
+    }
+
+    return ca;
+}

### 2018-07-30 simulation/SDL/src/main.c

@@ -1,0 +1,129 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <SDL2/SDL.h>
+
+#include "global.h"
+#include "automaton.h"
+
+int max_gens = 48;
+
+void rule_01(struct automaton *ca) {
+   
+    int *new = malloc(sizeof(int) * ca->len);
+    for (int i = 1; i < ca->len-1; i++) {
+        new[i] = (ca->cells[i-1] + ca->cells[i+1]) % 2;
+    }
+
+    new[0] = ca->cells[1];
+    new[ca->len-1] = ca->cells[ca->len-2];
+
+    ca->cells = new;
+
+}
+
+void render(SDL_Renderer *r, struct automaton *ca) {
+
+    /* draw to screen */
+    for (int i = 0; i < ca->len; i++) {
+        if (ca->cells[i] == ALIVE) {
+            SDL_SetRenderDrawColor(r, 255, 255, 255, SDL_ALPHA_OPAQUE);
+        } else {
+            SDL_SetRenderDrawColor(r, 0, 0, 0, SDL_ALPHA_OPAQUE);
+        }
+        SDL_RenderFillRect(r, &ca->rects[i]);
+        ca->rects[i].y += ca->rects[i].h;
+    }
+}
+
+int main() {
+
+    if (SDL_Init(SDL_INIT_VIDEO | SDL_INIT_AUDIO | SDL_INIT_EVENTS) != 0) {
+        fprintf(stderr, "Error initialising SDL2: %s\n", SDL_GetError());
+        exit(EXIT_FAILURE);
+    }
+
+    SDL_Window *window;
+    SDL_Renderer *renderer;
+    
+    window = SDL_CreateWindow(
+            "Cellular Automata",
+            SDL_WINDOWPOS_CENTERED,
+            SDL_WINDOWPOS_CENTERED,
+            WIN_WIDTH,
+            WIN_HEIGHT,
+            SDL_WINDOW_SHOWN
+    );
+
+    if (window == NULL) {
+        fprintf(stderr, "Could not create window: %s\n", SDL_GetError());
+        exit(EXIT_FAILURE);
+    }
+
+    renderer = SDL_CreateRenderer(
+            window,
+            -1,
+            SDL_RENDERER_ACCELERATED
+    );
+
+    if (renderer == NULL) {
+        fprintf(stderr, "Could not create renderer: %s\n", SDL_GetError());
[…]

### 2018-07-31 simulation/include/global.h
Large rule 1

@@ -1,8 +1,8 @@
 #ifndef _GLOBAL_H
 #define _GLOBAL_H
 
-#define WIN_WIDTH 640
-#define WIN_HEIGHT 480
+#define WIN_WIDTH 1024
+#define WIN_HEIGHT 768
 
 #define ALIVE 1
 #define DEAD 0

### 2018-07-31 simulation/src/automaton.c

@@ -4,7 +4,7 @@
 
 struct automaton *init_automaton(int len, simulate_fn func, int d) {
     struct automaton *ca = malloc(sizeof(struct automaton));
-    ca->cells = calloc(len, sizeof(int));
+    ca->cells = calloc(len * d, sizeof(int));
     
     ca->len = len;
     ca->dimension = d;

### 2018-07-31 simulation/src/main.c

@@ -1,5 +1,6 @@
 #include <stdio.h>
 #include <stdlib.h>
+#include <math.h>
 #include <SDL2/SDL.h>
 
 #include "global.h"
@@ -9,7 +10,9 @@ int max_gens = 48;
 
 void rule_01(struct automaton *ca) {
    
+    /* create new array to compute states without changing current ones */
     int *new = malloc(sizeof(int) * ca->len);
+
     for (int i = 1; i < ca->len-1; i++) {
         new[i] = (ca->cells[i-1] + ca->cells[i+1]) % 2;
     }
@@ -21,6 +24,30 @@ void rule_01(struct automaton *ca) {
 
 }
 
+/*
+void gol(struct automaton *ca) {
+
+    int *new = malloc(sizeof(int) * ca->len * ca->len);
+    for (int i = 0; i < ca->len; i++) {
+        for (int j = 0; j < ca->len; j++) {
+
+            int alive = 0;
+
+            alive += ca->cells[i-1][j-1];
+            alive += ca->cells[i-1][j];
+            alive += ca->cells[i-1][j+1];
+            alive += ca->cells[i][j-1];
+            alive += ca->cells[i][j+1];
+            alive += ca->cells[i+1][j-1];
+            alive += ca->cells[i+1][j];
+            alive += ca->cells[i+1][j+1];
+
+        }
+    }
+
+}
+*/
+
 void render(SDL_Renderer *r, struct automaton *ca) {
 
     /* draw to screen */
@@ -30,6 +57,7 @@ void render(SDL_Renderer *r, struct automaton *ca) {
         } else {
             SDL_SetRenderDrawColor(r, 0, 0, 0, SDL_ALPHA_OPAQUE);
         }
+
         SDL_RenderFillRect(r, &ca->rects[i]);
         ca->rects[i].y += ca->rects[i].h;
     }
@@ -75,9 +103,21 @@ int main() {
 
     int config[] = {
         0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
-        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,1,
[…]

### 2018-08-02 simulation/include/global.h
Updates to paper, simulation, and gitignore

@@ -1,7 +1,7 @@
 #ifndef _GLOBAL_H
 #define _GLOBAL_H
 
-#define WIN_WIDTH 1024
+#define WIN_WIDTH 1025
 #define WIN_HEIGHT 768
 
 #define ALIVE 1

### 2018-08-02 simulation/src/main.c

@@ -109,15 +109,32 @@ int main() {
         0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
         0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
         0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
+        0,0,0,0,0,0,0,0,    0,0,0,0,1,0,0,0,
         0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
+        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
+        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
+        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
+        1,1,1,1,1,1,1,1,    1,1,1,1,1,1,1,1,
+        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
+        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
+        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
+
         1,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
         0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
+        0,0,0,0,0,1,0,0,    0,0,0,0,0,0,0,0,
         0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
         0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
         0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
         0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
         0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
         0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
+        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
+        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
+        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
+        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
+        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
+        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
[…]

### 2018-08-03 simulation/src/main.c
Updates to paper and larger automata

@@ -2,11 +2,12 @@
 #include <stdlib.h>
 #include <math.h>
 #include <SDL2/SDL.h>
+#include <time.h>
 
 #include "global.h"
 #include "automaton.h"
 
-int max_gens = 48;
+int max_gens = 1024;
 
 void rule_01(struct automaton *ca) {
    
@@ -23,11 +24,10 @@ void rule_01(struct automaton *ca) {
     ca->cells = new;
 
 }
-
 /*
 void gol(struct automaton *ca) {
 
-    int *new = malloc(sizeof(int) * ca->len * ca->len);
+    int *new = malloc(sizeof(int) * ca->len);
     for (int i = 0; i < ca->len; i++) {
         for (int j = 0; j < ca->len; j++) {
 
@@ -81,6 +81,8 @@ int main() {
             WIN_HEIGHT,
             SDL_WINDOW_SHOWN
     );
+    int win_width, win_height;
+    SDL_GetWindowSize(window, &win_width, &win_height);
 
     if (window == NULL) {
         fprintf(stderr, "Could not create window: %s\n", SDL_GetError());
@@ -100,43 +102,17 @@ int main() {
 
     SDL_Event e;
     int is_running = 1;
+    int current_gen = 0;
 
-    int config[] = {
-        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
-        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
-        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
-        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
-        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
-        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
-        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
-        0,0,0,0,0,0,0,0,    0,0,0,0,1,0,0,0,
-        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
-        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
-        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
-        0,0,0,0,0,0,0,0,    0,0,0,0,0,0,0,0,
[…]

### 2018-08-03 paper/paper.tex
Large rule 1
Added tex file

@@
 
+\bibliographystyle{ieeetr}
+
 \usepackage[margin=1in]{geometry}
+\usepackage{graphicx}
+\usepackage{subfig}
 \usepackage{amsmath}
+\usepackage{url}
 
@@
     potential in the context of solving two specific problems, the Firing Squad
-    Synchronization Problem and the Majority Problem. Existing solutions to
-    these problems are explored, and their existing avenues for optimization are
+    Synchronisation Problem and the Majority Problem. Existing solutions to
+    these problems are explored, and their existing avenues for optimisation are
     discussed. In order to fully appreciate the complex structures that can
     arise from such simple beginnings, this project also presents software to
-    visualize and probe further into the nature of the automata highlighted.
+    visualise and probe further into the nature of the automata highlighted.
 \end{abstract}
@@
     Consider the cellular automaton defined by the simple rule:
-    $$ a_{i}^{t+1} = a_{i-1}^{t} + a_{i+1}^{t} \mod{2} $$
+    \begin{equation}
+        \label{rule90}
+        a_{i}^{t+1} = a_{i-1}^{t} + a_{i+1}^{t} \mod{2}
+    \end{equation}
+
     For any automaton, in each time step the rule is applied to all cells in the
@@
 
+    \begin{figure}[h]
+        \begin{center}
+            \includegraphics[width=4in]{rule1.png}
+            \caption{Pattern generated from applying Eq. 1 over 129 generations}
+            \label{fig:rule90}
+        \end{center}
+    \end{figure}
+
+    \subsection{Self-similarity}
+    Figure \ref{fig:rule90} shows a visualisation of Eq. \ref{rule90}. A cell is
+    filled white if it has the value 1, and black otherwise. Note that each new
+    ``line'' of the visualisation represents the automaton in the next
+    successive time-step from the one before; this automaton is one-dimensional,
+    and so we may represent it in its entirety as a single row of cells. This
+    pattern exhibits two important characteristics, the first of which is
+    ``self-similarity'', meaning that portions of the pattern generated are
+    indistinguishable from the whole when magnified. As Wolfram states, in this
+    way ``the pattern is therefore invariant under rescaling... and may be
+    characterised by a fractal dimension''\cite{WolframFractal}. One such
+    definition of this is the Hausdorff-Besicovitch dimension\cite{Hausdorff}.
+
+    As an example, suppose we have a cube with some mass, and we wish to find
+    out how the mass scales when we try to make the same cube out of smaller
+    copies of the original. Intuitively, one way to do this require eight
+    smaller cubes, each of whose side length is half that of the original cube.
+    So with the scaling factor of $\frac{1}{2}$, the mass of each smaller cube
+    is $\frac{1}{8} = (\frac{1}{2})^3$.
+    So we arrive at the equation $N = S^D$, where $N$ is the number of smaller
+    copies that can be stuck together to make the original, $S$ is the scaling
+    factor of these smaller copies, and $D$ is the dimension. In this example, we
+    compute that the dimension of a cube is 3, which is obviously correct.  Back
+    to the dimension of the Sierpinski Triangle, we can see that each larger
+    triangle is made up of three smaller triangles, each of whose side lengths
+    are half that of the larger triangle. So we calculate the dimension as:
+
+    \begin{equation}
+        \label{sierpinskiDim}
+        \begin{split}
+            \tfrac{1}{3} & = (\tfrac{1}{2})^D \\
+            \log{\tfrac{1}{3}} &= D \log{\tfrac{1}{2}} \\
+            D &= \tfrac{\log{3}}{\log{2}} \approx 1.585
+        \end{split}
+    \end{equation}
+
+    Many naturally-occurring systems exhibit fractal structures, from snowflakes
+    to pine cones to Romanesco broccoli, and raises the possibility that these
+    are generated through the evolution of some natural cellular automata or
+    similar processes.
+
+    \begin{figure}[h]%
+        \centering
+        \subfloat[Snowflake]{{\includegraphics[width=5cm]{snowflake.jpg}}}%
+        \qquad
+        \subfloat[Romanesco broccoli]{{\includegraphics[width=5cm]{romanesco_broccoli.jpg}}}%
+        \caption{Snowflakes and Romanesco broccoli exhibit fractal structures}%
+        \label{fig:nature_fractals}%
+    \end{figure}
+
+    \subsection{Self-organisation}
+    Second of the characteristics exhibited by this cellular automaton is
+    ``self-organisation'',
+
+\section{}
 \section{Examples}
@@
 \section{Computational Universality}
-\section{Firing Squad Synchronization Problem}
+\section{Firing Squad Synchronisation Problem}
 \section{Majority Problem}
 \section{}
+
+\bibliography{bibliography}
+
 \end{document}

### 2018-08-13 simulation/include/automaton.h
Updates to paper and automata structures

@@ -4,27 +4,33 @@
 #include <SDL2/SDL.h>
 
 struct automaton;
-typedef void (*simulate_fn)(struct automaton *);
+typedef void (*simulate_fn)(void *);
 
 enum direction {
     UP, RIGHT, DOWN, LEFT
 };
 
 struct automaton {
-    int *cells;
-    int len;
-    int dimension;
-    SDL_Rect *rects;
-    int cell_width;
-    int cell_height;
+    int         *cells;
+    int         len;
+    int         dimension;
     simulate_fn sim;
+    int         max_gens;
+    SDL_Rect    *rects;
+    int         cell_width;
+    int         cell_height;
 };
 
 struct ant {
-    int pos;
-    int direction;
+    int x, y;
+    int dx, dy;
 };
 
-struct automaton *init_automaton(int len, simulate_fn func, int d);
+struct langton_config {
+    struct ant *ant;
+    struct automaton *automaton;
+};
+
+struct automaton *init_automaton(int len, simulate_fn func, int d, int g);
 
 #endif

### 2018-08-13 simulation/include/global.h

@@ -13,7 +13,10 @@ enum states {
     EMPTY,
     HEAD,
     TAIL,
-    CONDUCTOR
+    CONDUCTOR,
+
+    /* Langton's Ant */
+    ANT
 };
 
 enum modes {

### 2018-08-13 simulation/src/automaton.c

@@ -3,15 +3,16 @@
 #include "global.h"
 #include "automaton.h"
 
-struct automaton *init_automaton(int len, simulate_fn func, int d) {
+struct automaton *init_automaton(int len, simulate_fn func, int d, int g) {
     struct automaton *ca = malloc(sizeof(struct automaton));
 
     int total_cells = (int) pow(len, d);
     ca->cells = calloc(total_cells, sizeof(int));
     
     ca->len = len;
-    ca->dimension = d;
     ca->sim = func;
+    ca->dimension = d;
+    ca->max_gens = g;
 
     ca->rects = malloc(sizeof(SDL_Rect) * total_cells);
 
@@ -47,8 +48,9 @@ struct automaton *init_automaton(int len, simulate_fn func, int d) {
  * 111 110 101 100 011 010 001 000
  *  0   1   0   1   1   0   1   0
  */
-void rule_90(struct automaton *ca) {
+void rule_90(void *data) {
 
+    struct automaton *ca = (struct automaton *) data;
     int total_cells = (int) pow(ca->len, ca->dimension);
    
     /* create new array to compute states without changing current ones */
@@ -70,8 +72,9 @@ void rule_90(struct automaton *ca) {
  * 111 110 101 100 011 010 001 000
  *  0   1   1   0   1   1   1   0
  */
-void rule_110(struct automaton *ca) {
+void rule_110(void *data) {
 
+    struct automaton *ca = (struct automaton *) data;
     int total_cells = (int) pow(ca->len, ca->dimension);
 
     /* create new array to compute states without changing current ones */
@@ -96,8 +99,9 @@ void rule_110(struct automaton *ca) {
     ca->cells = new;
 }
 
-void gol(struct automaton *ca) {
+void gol(void *data) {
 
+    struct automaton *ca = (struct automaton *) data;
     int total_cells = (int) pow(ca->len, ca->dimension);
 
     int *new = calloc(total_cells, sizeof(int));
@@ -162,8 +166,9 @@ void gol(struct automaton *ca) {
  * conductor    -> head iff 2 neighbours are heads
  *              -> conductor otherwise
  */
-void wireworld(struct automaton *ca) {
+void wireworld(void *data) {
 
+    struct automaton *ca = (struct automaton *) data;
     int total_cells = (int) pow(ca->len, ca->dimension);
 
     int *new = calloc(total_cells, sizeof(int));
[…]

### 2018-08-13 simulation/src/main.c

@@ -10,6 +10,12 @@
 #include "global.h"
 #include "automaton.h"
 
+/* TODO
+ *  for langton's ant, pass ant coords in void * param (along with ca)
+ *  render active->cells[ant] as an unused color e.g. red
+ *  add ant coords to struct automaton ?
+ */
+
 pthread_t threads[2];
 pthread_mutex_t mutex;
 
@@ -22,11 +28,11 @@ int mode = SIMULATE;    /* enable/disable drawing */
 int paintbrush;         /* the state with which to overwrite cell */
 struct automaton *active;
 
-extern void rule_90(struct automaton *);
-extern void rule_110(struct automaton *);
-extern void gol(struct automaton *);
-extern void wireworld(struct automaton *);
-extern void langton(struct automaton *);
+extern void rule_90(void *);
+extern void rule_110(void *);
+extern void gol(void *);
+extern void wireworld(void *);
+extern void langton(void *);
 
 void render(struct automaton *);
 void handle_input(void);
@@ -67,8 +73,6 @@ int main() {
         exit(EXIT_FAILURE);
     }
 
-    /* program setup */
-
     srand(time(NULL));
 
     int config[512];
@@ -80,14 +84,16 @@ int main() {
     struct automaton *rule90 = init_automaton(
             sizeof(config) / sizeof(int),
             &rule_90,
-            1
+            1,
+            1000
     );
     rule90->cells = config;
 
     struct automaton *rule110 = init_automaton(
             sizeof(config) / sizeof(int),
             &rule_110,
-            1
+            1,
+            1000
     );
     rule110->cells = config;
 
@@ -103,22 +109,25 @@ int main() {
     struct automaton *wires = init_automaton(
             64,
             &wireworld,
-            2
+            2,
+            1000
     );
     wires->cells = ww_conf;
 
     struct automaton *conway = init_automaton(
             256,
             &gol,
-            2
+            2,
+            1000
     );
     conway->cells = gol_conf;
 
-    int *lang = calloc(512*512, sizeof(int));
+    int *lang = calloc(128*128, sizeof(int));
     struct automaton *ant = init_automaton(
-            512,
+            128,
             &langton,
-            2
+            2,
+            1000
     );
     ant->cells = lang;
 
@@ -134,7 +143,7 @@ int main() {
         if (!paused) {
             /* update automaton */
             active->sim(active);
[…]

### 2018-08-14 paper/paper.tex
Fixed double free in simulation, updated paper

@@
 \usepackage{url}
+\usepackage{diagbox}
 
@@
     \begin{equation}
-        \label{rule90}
+        \label{eq:rule90}
         a_{i}^{t+1} = a_{i-1}^{t} + a_{i+1}^{t} \mod{2}
@@
         \begin{center}
-            \includegraphics[width=4in]{rule1.png}
+            \includegraphics[width=4in]{rule90.png}
             \caption{Pattern generated from applying Eq. 1 over 129 generations}
@@
     \subsection{Self-similarity}
-    Figure \ref{fig:rule90} shows a visualisation of Eq. \ref{rule90}. A cell is
+    Figure \ref{fig:rule90} shows a visualisation of Eq. \ref{eq:rule90}. A cell is
     filled white if it has the value 1, and black otherwise. Note that each new
@@
     way ``the pattern is therefore invariant under rescaling... and may be
-    characterised by a fractal dimension''\cite{WolframFractal}. One such
-    definition of this is the Hausdorff-Besicovitch dimension\cite{Hausdorff}.
+    characterised by a fractal dimension'' \cite{WolframFractal}. One such
+    definition of this is the Hausdorff-Besicovitch dimension \cite{Hausdorff}.
 
@@
         \begin{split}
-            \tfrac{1}{3} & = (\tfrac{1}{2})^D \\
+            \tfrac{1}{3} &= (\tfrac{1}{2})^D \\
             \log{\tfrac{1}{3}} &= D \log{\tfrac{1}{2}} \\
@@
     Second of the characteristics exhibited by this cellular automaton is
-    ``self-organisation'',
+    ``self-organisation'', whereby from an initial disorderly or chaotic state
+    there appear ordered structures seemingly spontaneously. This is illustrated
+    in Fig. \ref{fig:self-organisation} below: one can clearly see the emergence
+    of similar triangular structures in the cone snail's shell, appearing from a
+    chaotic background. This starting chaos is modelled in the cellular
+    automaton by randomly assigning each cell a value of 0 or 1.
 
-\section{}
+    \begin{figure}[h]%
+        \centering
+        \subfloat[Shell of a cone
+        snail]{{\includegraphics[width=2in]{cone_shell.jpg}}}%
+        \qquad
+        \subfloat[Rule 90 with random starting
+        configuration]{{\includegraphics[width=2in]{rule90_rand.png}}}%
+        \caption{Natural processes can exhibit structures found in automata}
+        \label{fig:self-organisation}
+    \end{figure}
+
+\section{Classifying automata}
+    In order to classify automata, we can take into account both the size of the
+    neighbourhood each cell considers going into the next generation, as well as
+    the number of possible states in which a cell may be. Call these $N$ and $Q$
+    respectively; in the case of Eq. \ref{eq:rule90}, $N=3$ and $Q=2$, and so the
+    total number of different ``transitions'' from one state to the next is $N^Q
+    = 8$. We may write out all of these transitions as follows:
+
+    \begin{table}[h]
+        \centering
+        \begin{tabular}{c|c|c|c|c|c|c|c}
+            111 & 110 & 101 & 100 & 011 & 010 & 001 & 000 \\
+            \hline
+            0   & 1   & 0   & 1   & 1   & 0   & 1   & 0
+        \end{tabular}
+    \end{table}
+
+    Reading the bottom row of this transition table, we get the binary number
+    $01011010_2 = 90_{10}$, allowing us to name this rule set Rule 90. Using
+    this convention, we may now construct our own automata based solely on the
+    knowledge of its name (at least for elementary cellular automata). For
+    example, take Rule 110, introduced by Wolfram 1983 \cite{WolframRule110}. We
+    first convert $110_{10}$ to binary, and then use this to fill in the table
+    as above:
+
+    \begin{table}[h]
+        \begin{minipage}{.5\linewidth}
+            \centering
+            \begin{tabular}{c|c|c|c|c|c|c|c}
+                111 & 110 & 101 & 100 & 011 & 010 & 001 & 000 \\
+                \hline
+                ? & ? & ? & ? & ? & ? & ? & ?
+            \end{tabular}
+        \end{minipage}%
+        $ \Rightarrow $
+        \begin{minipage}{.6\linewidth}
+            \begin{tabular}{c|c|c|c|c|c|c|c}
+                111 & 110 & 101 & 100 & 011 & 010 & 001 & 000 \\
+                \hline
+                0 & 1 & 1 & 0 & 1 & 1 & 1 & 0
+            \end{tabular}
+        \end{minipage}
+    \end{table}
+
+    To convert this to a concrete set of rules to simulate in a computer
+    programs (in a more elegant way than a collection of eight \texttt{if}
+    statements), we can put the information we have gathered so far in a
+    Karnaugh map to get a Boolean algebraic expression of the ruleset, as in the
+    table below. We use $A$ , $B$, and $C$ to denote $a_{i-1}$, $a_i$, and
+    $a_{i+1}$, at time $t$, respectively. From this we can reduce the ruleset to
+    the Boolean expression in Eq. \ref{eq:rule110} by observation.
+
+    \begin{table}[h]
+        %\caption{The Karnaugh map resulting from decoding Rule 110}
+        %\label{kmap}
+        \centering
+        \begin{tabular}{|c||c|c|c|c|}\hline
+            \diagbox{C}{AB} & 00 & 01 & 11 & 10 \\
+            \hline
+            0 & 0  & 1  & 1  & 0    \\
+            \hline
+            1 & 1  & 1  & 0  & 1    \\
+            \hline
+        \end{tabular}
+    \end{table}
+
+    \begin{equation}
+        \label{eq:rule110}
+        \begin{split}
+            B^{t+1} &= \overline{A} \cdot C + B \cdot \overline{C} + \overline{B} \cdot C \\
+            &= \overline{A} \cdot C + B \oplus C
+        \end{split}
+    \end{equation}
+
+    The resulting visualisation in Fig. \ref{fig:rule110} exhibits similar
+    characteristic of self-organisation, however it is hard to say whether it
+    shows self-similarity as well.
+
+    \begin{figure}[h]
+        \begin{center}
+            \includegraphics[width=4in]{rule110.png}
+            \caption{Rule 110 generated using Eq. \ref{eq:rule110}}
+            \label{fig:rule110}
+        \end{center}
+    \end{figure}
+
+    \subsection{Elementary cellular automata}
+    \subsection{Life-like cellular automata}
+    \subsection{Cyclic cellular automata}
+
 \section{Examples}
@@
 \section{Majority Problem}
-\section{}
 

### 2018-08-14 simulation/src/automaton.c

@@ -274,15 +274,15 @@ void langton(void *data) {
 
     int k;
 
-    switch (ca->cells[ant]) {
+    switch (offset) {
     case DEAD:
-        ca->cells[ant] = ALIVE;
+        ca->cells[offset] = ALIVE;
         k = dx;
         dx = -dy;
         dy = k;
         break;
     case ALIVE:
-        ca->cells[ant] = DEAD;
+        ca->cells[offset] = DEAD;
         k = dy;
         dy = -dx;
         dx = k;

### 2018-08-14 simulation/src/main.c

@@ -75,14 +75,13 @@ int main() {
 
     srand(time(NULL));
 
-    int config[512];
-    for (int i = 0; i < 512; i++) {
-        config[i] = 0;
+    int *config = calloc(256, sizeof(int));
+    for (int i = 0; i < 256; i++) {
+        config[i] = rand() % 2;
     }
-    config[511] = 1;
 
     struct automaton *rule90 = init_automaton(
-            sizeof(config) / sizeof(int),
+            256,
             &rule_90,
             1,
             1000
@@ -90,7 +89,7 @@ int main() {
     rule90->cells = config;
 
     struct automaton *rule110 = init_automaton(
-            sizeof(config) / sizeof(int),
+            256,
             &rule_110,
             1,
             1000
@@ -131,7 +130,7 @@ int main() {
     );
     ant->cells = lang;
 
-    active = ant;
+    active = rule90;
 
     while (running) {
 
@@ -143,7 +142,7 @@ int main() {
         if (!paused) {
             /* update automaton */
             active->sim(active);
-            SDL_Delay(200);
+            SDL_Delay(20);
         }
 
     }

### 2018-08-17 simulation/include/global.h
FPS => CPU usage now regulated

@@ -4,6 +4,8 @@
 #define WIN_WIDTH 512
 #define WIN_HEIGHT 512
 
+#define FRAME_RATE 60
+
 enum states {
     /* GoL */
     DEAD = 0,

### 2018-08-17 simulation/src/main.c

@@ -3,7 +3,6 @@
 #include <stdbool.h>
 #include <math.h>
 #include <time.h>
-#include <pthread.h>
 
 #include <SDL2/SDL.h>
 
@@ -16,8 +15,8 @@
  *  add ant coords to struct automaton ?
  */
 
-pthread_t threads[2];
-pthread_mutex_t mutex;
+/* milliseconds for each frame to take */
+const int FRAME_INTERVAL = 1 * 1000 / FRAME_RATE;
 
 SDL_Window *window;
 SDL_Renderer *renderer;
@@ -75,13 +74,13 @@ int main() {
 
     srand(time(NULL));
 
-    int *config = calloc(256, sizeof(int));
-    for (int i = 0; i < 256; i++) {
+    int *config = calloc(512, sizeof(int));
+    for (int i = 0; i < 512; i++) {
         config[i] = rand() % 2;
     }
 
     struct automaton *rule90 = init_automaton(
-            256,
+            512,
             &rule_90,
             1,
             1000
@@ -96,11 +95,6 @@ int main() {
     );
     rule110->cells = config;
 
-    int *gol_conf = calloc(256 * 256, sizeof(int));
-    gol_conf[32768] = gol_conf[32771] = gol_conf[33028] = gol_conf[33280] =
-        gol_conf[33284] = gol_conf[33537] = gol_conf[33538] = gol_conf[33539] =
-        gol_conf[33540] = ALIVE;
-
     int *ww_conf = calloc(64 * 64, sizeof(int));
     for (int i = 0; i < 64*64; i++)
         ww_conf[i] = EMPTY;
@@ -113,8 +107,9 @@ int main() {
     );
     wires->cells = ww_conf;
 
+    int *gol_conf = calloc(128 * 128, sizeof(int));
     struct automaton *conway = init_automaton(
-            256,
+            128,
             &gol,
             2,
             1000
@@ -130,7 +125,13 @@ int main() {
     );
     ant->cells = lang;
 
-    active = rule90;
+    active = wires;
+
+    unsigned int frame_start;
+    unsigned int frame_curr;
+    unsigned int wait_time;
+
+    frame_start = SDL_GetTicks();
 
     while (running) {
 
@@ -142,8 +143,16 @@ int main() {
         if (!paused) {
             /* update automaton */
             active->sim(active);
-            SDL_Delay(20);
         }
+
[…]

### 2018-08-17 simulation/src/main.c
Basic straight line drawing

@@ -9,6 +9,8 @@
 #include "global.h"
 #include "automaton.h"
 
+#define max(a,b) ((a) >= (b) ? a : b)
+
 /* TODO
  *  for langton's ant, pass ant coords in void * param (along with ca)
  *  render active->cells[ant] as an unused color e.g. red
@@ -238,6 +240,13 @@ void print_help(void) {
 
 void handle_input() {
 
+    /* for drawing */
+    static int x1, y1, x2, y2;
+
+    /* for computing equation of line */
+    float m;
+    float c;
+
     /* keep track of which cell to modify in DRAW mode */
     int n_i, n_j;
 
@@ -272,6 +281,10 @@ void handle_input() {
             break;
         case SDL_KEYDOWN:
             switch (e.key.keysym.sym) {
+            case SDLK_LSHIFT:
+                SDL_GetMouseState(&x1, &y1);
+                printf("Pressed at (%d,%d)\n", x1, y1);
+                break;
             case SDLK_h:
                 print_help();
                 break;
@@ -319,7 +332,47 @@ void handle_input() {
                 paintbrush = DEAD;
             }
             break;
+        case SDL_KEYUP:
+            switch (e.key.keysym.sym) {
+            case SDLK_LSHIFT:
+                SDL_GetMouseState(&x2, &y2);
+
+                if (x1 != x2) {
+                    m = (y2-y1) / (x2-x1);
+                    c = y1 - m*x1;
+
+                    int upper = max(x1,x2);
+                    int lower = (upper == x1) ? x2 : x1;
+                    float y;
+
+                    for (float i = lower; i < upper; i+=.5) {
+                        y = m * i + c;
+                        n_i = i / active->cell_width;
+                        n_j = y / active->cell_height;
+
+                        active->cells[n_j*active->len + n_i] = paintbrush;
+                    }
+                }
+
+                break;
+            }
+
         }
+    }
+
+    Uint8* keystate = SDL_GetKeyboardState(NULL);
+
+    /* continuous response keys */
+    if (keystate[SDL_SCANCODE_LSHIFT]) {
+    }
+
+    if(keystate[SDLK_RIGHT]) {
+    }
+
+    if(keystate[SDLK_UP]) {
+    }
+
+    if(keystate[SDLK_DOWN]) {
     }
 
 }

### 2018-08-17 simulation/src/main.c
Linear interpolation for lines

@@ -17,6 +17,16 @@
  *  add ant coords to struct automaton ?
  */
 
+float lerp(float start, float end, float t) {
+    return start + t * (end - start);
+}
+
+int diag(int x1, int y1, int x2, int y2) {
+    int dx = x2 - x1;
+    int dy = y2 - y1;
+    return max(abs(dx), abs(dy));
+}
+
 /* milliseconds for each frame to take */
 const int FRAME_INTERVAL = 1 * 1000 / FRAME_RATE;
 
@@ -283,7 +293,6 @@ void handle_input() {
             switch (e.key.keysym.sym) {
             case SDLK_LSHIFT:
                 SDL_GetMouseState(&x1, &y1);
-                printf("Pressed at (%d,%d)\n", x1, y1);
                 break;
             case SDLK_h:
                 print_help();
@@ -337,21 +346,19 @@ void handle_input() {
             case SDLK_LSHIFT:
                 SDL_GetMouseState(&x2, &y2);
 
-                if (x1 != x2) {
-                    m = (y2-y1) / (x2-x1);
-                    c = y1 - m*x1;
+                int ai, aj, bi, bj;
+                ai = x1 / active->cell_width;
+                aj = y1 / active->cell_height;
+                bi = x2 / active->cell_width;
+                bj = y2 / active->cell_height;
 
-                    int upper = max(x1,x2);
-                    int lower = (upper == x1) ? x2 : x1;
-                    float y;
+                float N = (float) diag(ai, aj, bi, bj);
 
-                    for (float i = lower; i < upper; i+=.5) {
-                        y = m * i + c;
-                        n_i = i / active->cell_width;
-                        n_j = y / active->cell_height;
-
-                        active->cells[n_j*active->len + n_i] = paintbrush;
-                    }
+                for (int i = 0; i <= N; i++) {
+                    float t = (N == 0) ? 0.f : i / N;
+                    n_i = round(lerp(ai, bi, t));
+                    n_j = round(lerp(aj, bj, t));
+                    active->cells[n_j*active->len + n_i] = paintbrush;
                 }
 
                 break;

### 2018-08-17 simulation/src/automaton.c
Controls and modified drawable states

@@ -205,8 +205,8 @@ void wireworld(void *data) {
                 heads += (ca->cells[curr + ca->len + 1] == HEAD);
 
             switch(ca->cells[curr]) {
-            case EMPTY:
-                new[curr] = EMPTY;
+            case DEAD:
+                new[curr] = DEAD;
                 break;
             case HEAD:
                 new[curr] = TAIL;
@@ -221,7 +221,7 @@ void wireworld(void *data) {
                     new[curr] = CONDUCTOR;
                 break;
             default:
-                new[curr] = EMPTY;
+                new[curr] = DEAD;
             }
 
         }

### 2018-08-17 simulation/src/main.c

@@ -250,12 +250,8 @@ void print_help(void) {
 
 void handle_input() {
 
-    /* for drawing */
+    /* for mouse coords */
     static int x1, y1, x2, y2;
-
-    /* for computing equation of line */
-    float m;
-    float c;
 
     /* keep track of which cell to modify in DRAW mode */
     int n_i, n_j;
@@ -318,16 +314,14 @@ void handle_input() {
                 paintbrush = ALIVE;
                 break;
             case SDLK_2:
-                paintbrush = EMPTY;
-                break;
-            case SDLK_3:
                 paintbrush = HEAD;
                 break;
+            case SDLK_3:
+                paintbrush = CONDUCTOR;
+                break;
             case SDLK_4:
-                paintbrush = TAIL;
                 break;
             case SDLK_5:
-                paintbrush = CONDUCTOR;
                 break;
             case SDLK_6:
                 break;
@@ -338,7 +332,7 @@ void handle_input() {
             case SDLK_9:
                 break;
             default:
-                paintbrush = DEAD;
+                paintbrush = ALIVE;
             }
             break;
         case SDL_KEYUP:

### 2018-08-17 simulation/include/automaton.h
Removed simulate mode: unnecessary

@@ -33,4 +33,10 @@ struct langton_config {
 
 struct automaton *init_automaton(int len, simulate_fn func, int d, int g);
 
+void rule_90(void *);
+void rule_110(void *);
+void gol(void *);
+void wireworld(void *);
+void langton(void *);
+
 #endif

### 2018-08-17 simulation/include/global.h

@@ -21,9 +21,4 @@ enum states {
     ANT
 };
 
-enum modes {
-    SIMULATE,
-    DRAW
-};
-
 #endif

### 2018-08-17 simulation/src/main.c

@@ -35,19 +35,18 @@ SDL_Renderer *renderer;
 
 int running = true;
 int paused = false;
-int mode = SIMULATE;    /* enable/disable drawing */
-int paintbrush;         /* the state with which to overwrite cell */
-struct automaton *active;
+int mode = DRAW;            /* enable/disable drawing */
+int paintbrush = ALIVE;     /* the state with which to overwrite cell */
 
-extern void rule_90(void *);
-extern void rule_110(void *);
-extern void gol(void *);
-extern void wireworld(void *);
-extern void langton(void *);
+struct automaton *active;
+struct automaton *rule90;
+struct automaton *rule110;
+struct automaton *conway;
+struct automaton *wires;
+struct automaton *lang;
 
 void render(struct automaton *);
 void handle_input(void);
-void print_help(void);
 
 int main() {
 
@@ -91,7 +90,7 @@ int main() {
         config[i] = rand() % 2;
     }
 
-    struct automaton *rule90 = init_automaton(
+    rule90 = init_automaton(
             512,
             &rule_90,
             1,
@@ -99,7 +98,7 @@ int main() {
     );
     rule90->cells = config;
 
-    struct automaton *rule110 = init_automaton(
+    rule110 = init_automaton(
             256,
             &rule_110,
             1,
@@ -111,7 +110,7 @@ int main() {
     for (int i = 0; i < 64*64; i++)
         ww_conf[i] = EMPTY;
 
-    struct automaton *wires = init_automaton(
+    wires = init_automaton(
             64,
             &wireworld,
             2,
@@ -120,7 +119,8 @@ int main() {
     wires->cells = ww_conf;
 
     int *gol_conf = calloc(128 * 128, sizeof(int));
-    struct automaton *conway = init_automaton(
+    
+    conway = init_automaton(
             128,
             &gol,
             2,
@@ -128,7 +128,7 @@ int main() {
     );
     conway->cells = gol_conf;
 
-    int *lang = calloc(128*128, sizeof(int));
+    lang = calloc(128*128, sizeof(int));
     struct automaton *ant = init_automaton(
             128,
             &langton,
@@ -136,13 +136,17 @@ int main() {
             1000
     );
     ant->cells = lang;
-
-    active = wires;
-
+    
+    /* fps regulation */
     unsigned int frame_start;
     unsigned int frame_curr;
     unsigned int wait_time;
 
+    /* simulation speed */
+    int ticks = 0;
+
[…]

### 2018-08-17 simulation/src/main.c
Can now change burshes again

@@ -17,16 +17,6 @@
  *  add ant coords to struct automaton ?
  */
 
-float lerp(float start, float end, float t) {
-    return start + t * (end - start);
-}
-
-int diag(int x1, int y1, int x2, int y2) {
-    int dx = x2 - x1;
-    int dy = y2 - y1;
-    return max(abs(dx), abs(dy));
-}
-
 /* milliseconds for each frame to take */
 const int FRAME_INTERVAL = 1 * 1000 / FRAME_RATE;
 
@@ -35,7 +25,6 @@ SDL_Renderer *renderer;
 
 int running = true;
 int paused = false;
-int mode = DRAW;            /* enable/disable drawing */
 int paintbrush = ALIVE;     /* the state with which to overwrite cell */
 
 struct automaton *active;
@@ -47,6 +36,9 @@ struct automaton *lang;
 
 void render(struct automaton *);
 void handle_input(void);
+
+float lerp(float, float, float);
+int diag(int, int, int, int);
 
 int main() {
 
@@ -304,7 +296,7 @@ void handle_input() {
                     active = conway;
                     printf("Welcome to the Game of Life!\n");
                 } else {
-                    paintbrush = ALIVE;
+                    paintbrush = DEAD;
                 }
                 break;
             case SDLK_2:
@@ -320,7 +312,7 @@ void handle_input() {
                     active = rule90;
                     printf("Welcome to Rule 90!\n");
                 } else {
-                    paintbrush = ALIVE;
+                    paintbrush = HEAD;
                 }
                 break;
             case SDLK_4:
@@ -328,7 +320,7 @@ void handle_input() {
                     active = rule110;
                     printf("Welcome to Rule 110!\n");
                 } else {
-                    paintbrush = ALIVE;
+                    paintbrush = CONDUCTOR;
                 }
                 break;
             case SDLK_5:
@@ -390,4 +382,16 @@ void handle_input() {
     if(keystate[SDLK_DOWN]) {
     }
 
+}
+
+/* linear interpolation to approximate a line on the grid */
+float lerp(float start, float end, float t) {
+    return start + t * (end - start);
+}
+
+/* calculate value for N when interpolating */
+int diag(int x1, int y1, int x2, int y2) {
+    int dx = x2 - x1;
+    int dy = y2 - y1;
+    return max(abs(dx), abs(dy));
 }

### 2018-08-18 simulation/include/automaton.h
Fixed pause issue with 1d automata and removed max_gens

@@ -31,7 +31,7 @@ struct langton_config {
     struct automaton *automaton;
 };
 
-struct automaton *init_automaton(int len, simulate_fn func, int d, int g);
+struct automaton *init_automaton(int len, simulate_fn func, int d);
 
 void rule_90(void *);
 void rule_110(void *);

### 2018-08-18 simulation/src/automaton.c

@@ -3,7 +3,7 @@
 #include "global.h"
 #include "automaton.h"
 
-struct automaton *init_automaton(int len, simulate_fn func, int d, int g) {
+struct automaton *init_automaton(int len, simulate_fn func, int d) {
     struct automaton *ca = malloc(sizeof(struct automaton));
 
     int total_cells = (int) pow(len, d);
@@ -12,7 +12,6 @@ struct automaton *init_automaton(int len, simulate_fn func, int d, int g) {
     ca->len = len;
     ca->sim = func;
     ca->dimension = d;
-    ca->max_gens = g;
 
     ca->rects = malloc(sizeof(SDL_Rect) * total_cells);
 
@@ -63,6 +62,10 @@ void rule_90(void *data) {
     new[0] = ca->cells[1];
     new[ca->len-1] = ca->cells[ca->len-2];
 
+    for (int i = 0; i < ca->len; i++) {
+        ca->rects[i].y += ca->rects[i].h;
+    }
+
     free(ca->cells);
     ca->cells = new;
 
@@ -94,6 +97,10 @@ void rule_110(void *data) {
 
     new[0] = result;
     new[ca->len-1] = ca->cells[ca->len-1];
+
+    for (int i = 0; i < ca->len; i++) {
+        ca->rects[i].y += ca->rects[i].h;
+    }
 
     free(ca->cells);
     ca->cells = new;

### 2018-08-18 simulation/src/main.c

@@ -79,22 +79,21 @@ int main() {
 
     int *config = calloc(512, sizeof(int));
     for (int i = 0; i < 512; i++) {
-        config[i] = rand() % 2;
+        //config[i] = rand() % 2;
     }
+    config[255] = 1;
 
     rule90 = init_automaton(
             512,
             &rule_90,
-            1,
-            1000
+            1
     );
     rule90->cells = config;
 
     rule110 = init_automaton(
             256,
             &rule_110,
-            1,
-            1000
+            1
     );
     rule110->cells = config;
 
@@ -105,8 +104,7 @@ int main() {
     wires = init_automaton(
             64,
             &wireworld,
-            2,
-            1000
+            2
     );
     wires->cells = ww_conf;
 
@@ -115,8 +113,7 @@ int main() {
     conway = init_automaton(
             128,
             &gol,
-            2,
-            1000
+            2
     );
     conway->cells = gol_conf;
 
@@ -124,8 +121,7 @@ int main() {
     struct automaton *ant = init_automaton(
             128,
             &langton,
-            2,
-            1000
+            2
     );
     ant->cells = lang;
     
@@ -138,7 +134,7 @@ int main() {
     int ticks = 0;
 
     /* main program execution */
-    active = conway;
+    active = rule110;
     frame_start = SDL_GetTicks();
 
     while (running) {
@@ -189,7 +185,6 @@ void render(struct automaton *ca) {
             }
 
             SDL_RenderFillRect(renderer, &ca->rects[i]);
-            ca->rects[i].y += ca->rects[i].h;
         }
 
     } else {

### 2018-08-18 simulation/include/automaton.h
Added Rule 30. Replaced rules with bitwise ops

@@ -33,6 +33,7 @@ struct langton_config {
 
 struct automaton *init_automaton(int len, simulate_fn func, int d);
 
+void rule_30(void *);
 void rule_90(void *);
 void rule_110(void *);
 void gol(void *);

### 2018-08-18 simulation/src/automaton.c

@@ -45,6 +45,39 @@ struct automaton *init_automaton(int len, simulate_fn func, int d) {
 
 /**
  * 111 110 101 100 011 010 001 000
+ *  0   0   0   1   1   1   1   0
+ */
+void rule_30(void *data) {
+
+    struct automaton *ca = (struct automaton *) data;
+    int total_cells = (int) pow(ca->len, ca->dimension);
+   
+    /* create new array to compute states without changing current ones */
+    int *new = malloc(sizeof(int) * total_cells);
+
+    for (int i = 1; i < ca->len-1; i++) {
+        int a, b, c;
+        a = ca->cells[i-1];
+        b = ca->cells[i];
+        c = ca->cells[i+1];
+        
+        new[i] = (~a & (b | c)) | (a & ~b & ~c);
+    }
+
+    new[0] = ca->cells[0] | ca->cells[1];
+    new[ca->len-1] = ca->cells[ca->len-2] ^ ca->cells[ca->len-1];
+
+    for (int i = 0; i < ca->len; i++) {
+        ca->rects[i].y += ca->rects[i].h;
+    }
+
+    free(ca->cells);
+    ca->cells = new;
+
+}
+
+/**
+ * 111 110 101 100 011 010 001 000
  *  0   1   0   1   1   0   1   0
  */
 void rule_90(void *data) {
@@ -56,7 +89,7 @@ void rule_90(void *data) {
     int *new = malloc(sizeof(int) * total_cells);
 
     for (int i = 1; i < ca->len-1; i++) {
-        new[i] = (ca->cells[i-1] + ca->cells[i+1]) % 2;
+        new[i] = ca->cells[i-1] ^ ca->cells[i+1];
     }
 
     new[0] = ca->cells[1];
@@ -85,17 +118,15 @@ void rule_110(void *data) {
     int result;
 
     for (int i = 1; i < ca->len-1; i++) {
-        result = (
-                (ca->cells[i] + ca->cells[i+1]) % 2) +
-                !ca->cells[i-1] * ca->cells[i+1];
-        result = (result > 0) ? 1 : 0;
-        new[i] = result;
+        int a, b, c;
+        a = ca->cells[i-1];
+        b = ca->cells[i];
+        c = ca->cells[i+1];
+
+        new[i] = (~a & c) | (b ^ c);
     }
 
-    result = ((ca->cells[0] + ca->cells[1]) % 2) + ca->cells[1];
[…]

### 2018-08-18 simulation/src/main.c

@@ -27,12 +27,14 @@ int running = true;
 int paused = false;
 int paintbrush = ALIVE;     /* the state with which to overwrite cell */
 
-struct automaton *active;
+struct automaton *rule30;
 struct automaton *rule90;
 struct automaton *rule110;
 struct automaton *conway;
 struct automaton *wires;
 struct automaton *lang;
+
+struct automaton *active;
 
 void render(struct automaton *);
 void handle_input(void);
@@ -78,10 +80,14 @@ int main() {
     srand(time(NULL));
 
     int *config = calloc(512, sizeof(int));
-    for (int i = 0; i < 512; i++) {
-        //config[i] = rand() % 2;
-    }
-    config[255] = 1;
+    config[256] = 1;
+
+    rule30 = init_automaton(
+            512,
+            &rule_30,
+            1
+    );
+    rule30->cells = config;
 
     rule90 = init_automaton(
             512,
@@ -134,7 +140,7 @@ int main() {
     int ticks = 0;
 
     /* main program execution */
-    active = rule110;
+    active = rule30;
     frame_start = SDL_GetTicks();
 
     while (running) {
@@ -158,8 +164,10 @@ int main() {
 
         if (frame_curr < frame_start + FRAME_INTERVAL) {
             wait_time = frame_start + FRAME_INTERVAL - frame_curr;
-            SDL_Delay(wait_time);
+            // SDL_Delay(wait_time);
         }
+
+        SDL_Delay(4);
 
         frame_start = frame_curr;
 

### 2018-08-18 simulation/include/automaton.h
Brian's Brain and various polish

@@ -36,7 +36,8 @@ struct automaton *init_automaton(int len, simulate_fn func, int d);
 void rule_30(void *);
 void rule_90(void *);
 void rule_110(void *);
-void gol(void *);
+void brians_brain(void *);
+void game_of_life(void *);
 void wireworld(void *);
 void langton(void *);
 

### 2018-08-18 simulation/include/global.h

@@ -7,12 +7,16 @@
 #define FRAME_RATE 60
 
 enum states {
-    /* GoL */
+
+    /* Game of Life */
     DEAD = 0,
     ALIVE = 1,
+
+    /* BRIAN'S BRAIN */
+    FIRING,
+    DYING,
     
     /* WireWorld */
-    EMPTY,
     HEAD,
     TAIL,
     CONDUCTOR,

### 2018-08-18 simulation/src/automaton.c

@@ -115,7 +115,6 @@ void rule_110(void *data) {
 
     /* create new array to compute states without changing current ones */
     int *new = malloc(sizeof(int) * total_cells);
-    int result;
 
     for (int i = 1; i < ca->len-1; i++) {
         int a, b, c;
@@ -137,7 +136,65 @@ void rule_110(void *data) {
     ca->cells = new;
 }
 
-void gol(void *data) {
+void brians_brain(void *data) {
+
+    struct automaton *ca = (struct automaton *) data;
+    int total_cells = (int) pow(ca->len, ca->dimension);
+
+    int *new = calloc(total_cells, sizeof(int));
+    memcpy(new, ca->cells, sizeof(int) * total_cells);
+
+    for (int i = 0; i < ca->len; i++) {
+        for (int  j = 0; j < ca->len; j++) {
+
+            int curr = i * ca->len + j;
+            int firing = 0;
+
+            if (ca->cells[curr] == FIRING)
+                new[curr] = DYING;
+
+            if (ca->cells[curr] == DYING)
+                new[curr] = DEAD;
+
+            if (ca->cells[curr] == DEAD) {
+                if (i > 0 && j > 0)
+                    firing += (ca->cells[curr - ca->len - 1] == FIRING);
+
+                if (i > 0)
+                    firing += (ca->cells[curr - ca->len] == FIRING);
+
+                if (i > 0 && j < ca->len-1)
+                    firing += (ca->cells[curr - ca->len + 1] == FIRING);
+
+                if (j > 0)
+                    firing += (ca->cells[curr - 1] == FIRING);
+
+                if (j < ca->len-1)
+                    firing += (ca->cells[curr + 1] == FIRING);
+
+                if (i < ca->len-1 && j > 0)
+                    firing += (ca->cells[curr + ca->len - 1] == FIRING);
+
+                if (i < ca->len-1)
+                    firing += (ca->cells[curr + ca->len] == FIRING);
+
+                if (i < ca->len-1 && j < ca->len-1)
+                    firing += (ca->cells[curr + ca->len + 1] == FIRING);
+
+                if (firing == 2)
+                    new[curr] = FIRING;
+
+            }
+
+        }
+    }
+
+    free(ca->cells);
+    ca->cells = new;
+
+}
+
+void game_of_life(void *data) {
 
     struct automaton *ca = (struct automaton *) data;
     int total_cells = (int) pow(ca->len, ca->dimension);

### 2018-08-18 simulation/src/main.c

@@ -31,6 +31,7 @@ struct automaton *rule30;
 struct automaton *rule90;
 struct automaton *rule110;
 struct automaton *conway;
+struct automaton *brian;
 struct automaton *wires;
 struct automaton *lang;
 
@@ -104,9 +105,6 @@ int main() {
     rule110->cells = config;
 
     int *ww_conf = calloc(64 * 64, sizeof(int));
-    for (int i = 0; i < 64*64; i++)
-        ww_conf[i] = EMPTY;
-
     wires = init_automaton(
             64,
             &wireworld,
@@ -115,13 +113,20 @@ int main() {
     wires->cells = ww_conf;
 
     int *gol_conf = calloc(128 * 128, sizeof(int));
-    
     conway = init_automaton(
             128,
-            &gol,
+            &game_of_life,
             2
     );
     conway->cells = gol_conf;
+
+    int *brain_conf = calloc(128*128, sizeof(int));
+    brian = init_automaton(
+            128,
+            &brians_brain,
+            2
+    );
+    brian->cells = brain_conf;
 
     lang = calloc(128*128, sizeof(int));
     struct automaton *ant = init_automaton(
@@ -140,7 +145,7 @@ int main() {
     int ticks = 0;
 
     /* main program execution */
-    active = rule30;
+    active = brian;
     frame_start = SDL_GetTicks();
 
     while (running) {
@@ -151,7 +156,7 @@ int main() {
         render(active);
 
         if (!paused) {
-            if (ticks > 5) {
+            if (ticks > 3) {
                 /* update automaton */
                 active->sim(active);
                 ticks = 0;
@@ -164,10 +169,8 @@ int main() {
 
         if (frame_curr < frame_start + FRAME_INTERVAL) {
             wait_time = frame_start + FRAME_INTERVAL - frame_curr;
-            // SDL_Delay(wait_time);
+            SDL_Delay(wait_time);
         }
-
-        SDL_Delay(4);
 
         frame_start = frame_curr;
 
@@ -209,8 +212,11 @@ void render(struct automaton *ca) {
                 case ALIVE:
                     SDL_SetRenderDrawColor(renderer, 255, 255, 255, SDL_ALPHA_OPAQUE);
                     break;
-                case EMPTY:
-                    SDL_SetRenderDrawColor(renderer, 0, 0, 0, SDL_ALPHA_OPAQUE);
+                case FIRING:
[…]

### 2018-08-18 simulation/src/main.c
Step forward

@@ -23,8 +23,9 @@ const int FRAME_INTERVAL = 1 * 1000 / FRAME_RATE;
 SDL_Window *window;
 SDL_Renderer *renderer;
 
-int running = true;
-int paused = false;
+bool running = true;
+bool paused = false;
+bool step = false;
 int paintbrush = ALIVE;     /* the state with which to overwrite cell */
 
 struct automaton *rule30;
@@ -104,9 +105,9 @@ int main() {
     );
     rule110->cells = config;
 
-    int *ww_conf = calloc(64 * 64, sizeof(int));
+    int *ww_conf = calloc(128*128, sizeof(int));
     wires = init_automaton(
-            64,
+            128,
             &wireworld,
             2
     );
@@ -162,6 +163,11 @@ int main() {
                 ticks = 0;
             }
             ticks++;
+        } else {
+            if (step) {
+                active->sim(active);
+                step = !step;
+            }
         }
 
         /* fix framerate at FRAME_RATE fps */
@@ -291,6 +297,9 @@ void handle_input() {
             case SDLK_LCTRL:
                 ctrl = true;
                 break;
+            case SDLK_RIGHT:
+                step = true;
+                break;
             case SDLK_p:
                 if (paused)
                     printf("Resuming...\n");
@@ -336,7 +345,7 @@ void handle_input() {
             case SDLK_3:
                 if (ctrl) {
                     active = brian;
-                    printf("Welcome to Rule 90!\n");
+                    printf("Welcome to Brian's Brain!\n");
                 } else {
                     paintbrush = HEAD;
                 }

### 2018-08-19 simulation/include/automaton.h
Langton's Ant finished. Generation conuter implemented

@@ -1,6 +1,7 @@
 #ifndef _AUTOMATON_H
 #define _AUTOMATON_H
 
+#include <stdbool.h>
 #include <SDL2/SDL.h>
 
 struct automaton;
@@ -8,6 +9,11 @@ typedef void (*simulate_fn)(void *);
 
 enum direction {
     UP, RIGHT, DOWN, LEFT
+};
+
+struct ant {
+    int x, y;
+    int dx, dy;
 };
 
 struct automaton {
@@ -19,19 +25,12 @@ struct automaton {
     SDL_Rect    *rects;
     int         cell_width;
     int         cell_height;
-};
-
-struct ant {
-    int x, y;
-    int dx, dy;
-};
-
-struct langton_config {
     struct ant *ant;
-    struct automaton *automaton;
+    bool        is_langton;
 };
 
 struct automaton *init_automaton(int len, simulate_fn func, int d);
+struct ant *init_ant(int x, int y, int dx, int dy, struct automaton *);
 
 void rule_30(void *);
 void rule_90(void *);
@@ -39,6 +38,6 @@ void rule_110(void *);
 void brians_brain(void *);
 void game_of_life(void *);
 void wireworld(void *);
-void langton(void *);
+void langtons_ant(void *);
 
 #endif

### 2018-08-19 simulation/src/automaton.c

@@ -40,7 +40,22 @@ struct automaton *init_automaton(int len, simulate_fn func, int d) {
 
     }
 
+    ca->is_langton = false;
+
     return ca;
+}
+
+struct ant *init_ant(int x, int y, int dx, int dy, struct automaton *ca) {
+    struct ant *ant = malloc(sizeof(struct ant));
+    ant->x = x;
+    ant->y = y;
+    ant->dx = dx;
+    ant->dy = dy;
+
+    ca->ant = ant;
+    ca->is_langton = true;
+
+    return ant;
 }
 
 /**
@@ -332,58 +347,31 @@ void wireworld(void *data) {
  * white -> black, turn right
  * move ant forward
  */
-void langton(void *data) {
+void langtons_ant(void *data) {
 
-    struct langton_config *lc = (struct langton_config *) data;
+    struct automaton *ca = (struct automaton *) data;
 
-    struct ant *ant = lc->ant;
-    struct automaton *ca = lc->automaton;
-
-    int x = ant->x;
-    int y = ant->y;
-    int dx = ant->dx;
-    int dy = ant->dy;
+    int x = ca->ant->x;
+    int y = ca->ant->y;
+    int dx = ca->ant->dx;
+    int dy = ca->ant->dy;
 
     int offset = y * ca->len + x;
 
-    // int total_cells = (int) pow(ca->len, ca->dimension);
-
-    // int *new = calloc(total_cells, sizeof(int));
-    // memcpy(new, ca->cells, sizeof(int) * total_cells);
-
-    /*
-     *  a.b = |a||b| cos theta
-     *  a.b = 0 (orthogonal)
-     *
-     *  we have a, want b
-     *  ax.bx + ay.by = 0
-     *  bx = -(ay.by)/ax
-     *
-     *  |b| = 1 = sqrt(b.b) = pow(bx,2)+pow(by,2)
-     *  bx = sqrt(1 - pow(by,2))
-     *
-     *  [ALGEBRA]
-     *
-     *  by = +/- sqrt(1/(1+pow(ay,2)/pow(ax,2)))
-     */
-
-    int k;
-
-    switch (offset) {
+    switch (ca->cells[offset]) {
     case DEAD:
         ca->cells[offset] = ALIVE;
-        k = dx;
-        dx = -dy;
-        dy = k;
+        ca->ant->dx = dy;
+        ca->ant->dy = -dx;
         break;
     case ALIVE:
[…]

### 2018-08-19 simulation/src/main.c

@@ -5,9 +5,12 @@
 #include <time.h>
 
 #include <SDL2/SDL.h>
+#include <SDL2/SDL_ttf.h>
 
 #include "global.h"
 #include "automaton.h"
+
+#define TXT_WIDTH 9
 
 #define max(a,b) ((a) >= (b) ? a : b)
 
@@ -23,10 +26,18 @@ const int FRAME_INTERVAL = 1 * 1000 / FRAME_RATE;
 SDL_Window *window;
 SDL_Renderer *renderer;
 
+SDL_Surface *surface;
+SDL_Texture *texture;
+SDL_Rect msg_rect;
+
 bool running = true;
 bool paused = false;
 bool step = false;
 int paintbrush = ALIVE;     /* the state with which to overwrite cell */
+
+/* simulation speed */
+int speed = 0;
+int gens = 0;
 
 struct automaton *rule30;
 struct automaton *rule90;
@@ -34,7 +45,7 @@ struct automaton *rule110;
 struct automaton *conway;
 struct automaton *brian;
 struct automaton *wires;
-struct automaton *lang;
+struct automaton *langton;
 
 struct automaton *active;
 
@@ -79,93 +90,100 @@ int main() {
         exit(EXIT_FAILURE);
     }
 
-    srand(time(NULL));
+    if(TTF_Init() == -1) {
+        fprintf(stderr, "Error on TTF_Init(): %s\n", TTF_GetError());
+        exit(EXIT_FAILURE);
+    }
 
-    int *config = calloc(512, sizeof(int));
-    config[256] = 1;
+    int ticks = 0;
+    char gen_str[12];
+    int gen_len = 1;
+    int last_gen_len = gen_len;
 
-    rule30 = init_automaton(
-            512,
-            &rule_30,
-            1
-    );
+    /* text setup */
+    TTF_Font *font = TTF_OpenFont("consolas.ttf", 144);
+    if(!font) {
+        fprintf(stderr, "TTF_OpenFont: %s\n", TTF_GetError());
+    }
+
+    SDL_Colour white = {255, 255, 255};
+
+    msg_rect.w = 1 * TXT_WIDTH;
+    msg_rect.h = 18;
+    msg_rect.x = WIN_WIDTH - msg_rect.w;
+    msg_rect.y = 0;
+
+    /* automata initialisation */
+    int *config = calloc(128*128, sizeof(int));
+
+    rule30 = init_automaton(512, &rule_30, 1);
     rule30->cells = config;
 
-    rule90 = init_automaton(
[…]

### 2018-08-19 simulation/src/main.c
Reshuffle and clean

@@ -10,15 +10,7 @@
 #include "global.h"
 #include "automaton.h"
 
-#define TXT_WIDTH 9
-
 #define max(a,b) ((a) >= (b) ? a : b)
-
-/* TODO
- *  for langton's ant, pass ant coords in void * param (along with ca)
- *  render active->cells[ant] as an unused color e.g. red
- *  add ant coords to struct automaton ?
- */
 
 /* milliseconds for each frame to take */
 const int FRAME_INTERVAL = 1 * 1000 / FRAME_RATE;

### 2018-08-19 paper/paper.tex
More elementary CA. Updated paper

@@
     finite collection of ``cells'', each in one of a finite number of states.
-    The state of each of these cells may evolve over the discrete progression of
-    time in discrete steps, and may do so according to a deterministic set of
-    rules specifying the state to which each cell is to progress, taking into
-    account the states of cells in its neighbourhood. Their inherently discrete
-    nature allows for strong analogies to be made with digital computers, and
-    gifts them the ability to simulate digital processes and the potential to
-    solve problems in this area.
+    The state of each of these cells may evolve over the progression of time in
+    discrete steps, and may do so according to a deterministic set of rules
+    based on the states of neighbouring cells. Their inherently discrete nature
+    allows for strong analogies to be made with digital computers, and gifts
+    them the ability to simulate digital processes and the potential to solve
+    problems in this area.
 
@@
     Reading the bottom row of this transition table, we get the binary number
-    $01011010_2 = 90_{10}$, allowing us to name this rule set Rule 90. Using
-    this convention, we may now construct our own automata based solely on the
-    knowledge of its name (at least for elementary cellular automata). For
-    example, take Rule 110, introduced by Wolfram 1983 \cite{WolframRule110}. We
-    first convert $110_{10}$ to binary, and then use this to fill in the table
+    $01011010_{\text{bin}} = 90_{\text{dec}}$, allowing us to name this rule set
+    Rule 90. Using this convention, we may now construct our own automata based
+    solely on the knowledge of its name (at least for elementary cellular
+    automata). For example, take Rule 110, introduced by Wolfram 1983
+    \cite{WolframRule110}. We first convert $110_{10}$ to binary, and then use
+    this to fill in the table
     as above:
@@
     \subsection{Elementary cellular automata}
+    Elementary cellular automata are the simplest class of one-dimensional
+    cellular automata; each cell may be in one of two states, 0 or 1, and rule
+    sets only take into account the cell itself and its immediate neighbours
+    (the nearest cell left and right of the cell being computed). Since there
+    are a total of $2^3=8$ combinations of values for a neighbourhood of cells,
+    and each of these neighbourhoods causes a cell to move to one of two states
+    in the next generation, this means there are $2^8=256$ elementary cellular
+    automata. Their evolution is often illustrated by each progressive ``new
+    line'' showing the next generation of the automaton, as caused by applying
+    the rule set to the current generation.
+
+    Two technically different automata may give rise to similar patterns, and
+    indeed many automata are equivalent to each other thanks to simple
+    transformations of theirunerlying geometry. The first of these
+    transformations is reflection in the vertical axis; the result of applying
+    this is known as the mirror rule \cite{elementary}. For example, take Rule
+    30. Each time some neighbourhood produces a live cell in Rule 30, let's take
+    this neighbourhood and reflect it in the vertical plane, giving the table
+    below:
+
+    \begin{table}[h]
+        \centering
+        \begin{tabular}{|c|c|c|c|c|c|c|c|}\hline
+            111 & 110 & 101 & 100 & 011 & 010 & 001 & 000   \\
+            \hline
+            0 & 0 & 0 & 1 & 1 & 1 & 1 & 0   \\
+            \hline
+            0 & 1 & 0 & 1 & 0 & 1 & 1 & 0   \\
+            \hline
+        \end{tabular}
+    \end{table}
+
+    The first line of the table shows the ruleset for Rule 30, while the line
+    below gives that of our new ruleset, made by applying the mirror rule to
+    Rule 30. Thus we get Rule 30's mirror rule Rule $01010110_{\text{bin}}$ =
+    Rule $86_{\text{dec}}$. Rules which generate the same ruleset under
+    mirroring are called amphichiral, and of the 256 elementary cellular
+    automata, 64 are so.
+
+    The second of these transformations is to exchange the roles of 0s and 1s in
+    the ruleset definition, and doing so is applying the complement rule
+    \cite{elementary}.  Again applying this to Rule 30, we get the table below
+    (note that the roles of 0s and 1s are exchanged in both the top and bottom
+    rows):
+
+    \begin{table}[h]
+        \centering
+        \begin{tabular}{|c|c|c|c|c|c|c|c|}\hline
+            000 & 001 & 010 & 011 & 100 & 101 & 110 & 111   \\
+            \hline
+            1 & 1 & 1 & 0 & 0 & 0 & 0 & 1   \\
+            \hline
+        \end{tabular}
+    \end{table}
+
+    So we arrive at Rule $10000111_{\text{bin}} = 135_{\text{dec}}$ as the
+    complement of Rule 30. There are 16 rules which are the same as their
+    complement.
+    Both of these rules can be applied at once to a ruleset, and the result is
+    Rule 149 when applied to Rule 30. There are 16 rules which are the same as
+    their mirrored complementary rules. There are 88 rules which are
+    inequivalent under these transformations \cite{elementary}.
+
+    \subsection{Wolfram classes}
+    Wolfram gives the following classes for categorising the behaviours of
+    different automata:
+
+    \begin{itemize}
+        \item Class One: rapidly converge to a uniform state e.g. Rule 0, 232
+        \item Class Two: rapidly converge to a repetitive or stable state e.g.
+            Rule 250
+        \item Class Three: appear to remain in a random state e.g. Rule 150, 182
+        \item Class Four: form areas of repetitive or stable states, but also
+            structures which interact in complicated ways e.g. Rule 54
+    \end{itemize}
+
+    Two of these 256 rules are particularly special, namely Rule 54 and 110,
+    both of which are Class 4 elementary cellular automata. Rule 110 has been
+    proven to be universal, or Turing complete, meaning it is theorectically
+    able to simulate any function. Meanwhile, it is so far unknown whether Rule
+    54 is capable of universal computation -- interacting structures form, but
+    it remains to been seen whether structures useful for computation can be
+    formed. Computational universality is discussed more in-depth later on.
+
+    \subsection{Two dimensional cellular automata}
+    Two dimensional cellular automata become a bit more interesting than their
+    dimensionally-deficient brethren. Rather than having a single row of cells
+    to consider at each generation, now a two-dimensional plane of cells is used
+    to compute the evolution of the automaton. Whereas elementary automata
+    looked only at the neighbourhood of $a_{i-1}, a_i,$ and $a_{i+1}$ when
+    applying the ruleset, two-dimensional automata have a wider choice. Two of
+    the most common types of neighbourhood to consider are the Moore
+    neighbourhood and the von Neumann neighbourhood. The former includes the 8
+    cells directly surrounding the current cell being considered, while the
+    latter only focuses on the four orthogonally ajacent cells. Ocassionally,
+    one may use the extended von Neumann neighbourhood.
+
     \subsection{Life-like cellular automata}

### 2018-08-19 simulation/include/automaton.h

@@ -7,10 +7,6 @@
 struct automaton;
 typedef void (*simulate_fn)(void *);
 
-enum direction {
-    UP, RIGHT, DOWN, LEFT
-};
-
 struct ant {
     int x, y;
     int dx, dy;
@@ -32,9 +28,16 @@ struct automaton {
 struct automaton *init_automaton(int len, simulate_fn func, int d);
 struct ant *init_ant(int x, int y, int dx, int dy, struct automaton *);
 
+/* elementary cellular automata */
 void rule_30(void *);
+void rule_54(void *);
 void rule_90(void *);
 void rule_110(void *);
+void rule_150(void *);
+void rule_182(void *);
+void rule_232(void *);
+void rule_250(void *);
+
 void brians_brain(void *);
 void game_of_life(void *);
 void wireworld(void *);

### 2018-08-19 simulation/include/global.h

@@ -6,6 +6,8 @@
 
 #define FRAME_RATE 60
 
+#define TXT_WIDTH 9
+
 enum states {
 
     /* Game of Life */

### 2018-08-19 simulation/src/automaton.c

@@ -88,8 +88,40 @@ void rule_30(void *data) {
 
     free(ca->cells);
     ca->cells = new;
-
 }
+
+/**
+ * 111 110 101 100 011 010 001 000
+ *  0   0   1   1   0   1   1   0
+ */
+void rule_54(void *data) {
+
+    struct automaton *ca = (struct automaton *) data;
+    int total_cells = (int) pow(ca->len, ca->dimension);
+   
+    /* create new array to compute states without changing current ones */
+    int *new = malloc(sizeof(int) * total_cells);
+
+    for (int i = 1; i < ca->len-1; i++) {
+        int a, b, c;
+        a = ca->cells[i-1];
+        b = ca->cells[i];
+        c = ca->cells[i+1];
+        
+        new[i] = (~a & b & ~c) | (~b & (a | c));
+    }
+
+    new[0] = ca->cells[0] ^ ca->cells[1];
+    new[ca->len-1] = ca->cells[ca->len-2] ^ ca->cells[ca->len-1];
+
+    for (int i = 0; i < ca->len; i++) {
+        ca->rects[i].y += ca->rects[i].h;
+    }
+
+    free(ca->cells);
+    ca->cells = new;
+}
+
 
 /**
  * 111 110 101 100 011 010 001 000
@@ -116,7 +148,6 @@ void rule_90(void *data) {
 
     free(ca->cells);
     ca->cells = new;
-
 }
 
 /**
@@ -142,6 +173,113 @@ void rule_110(void *data) {
 
     new[0] = ca->cells[1] | (ca->cells[0] ^ ca->cells[1]);
     new[ca->len-1] = ca->cells[ca->len-1];
+
+    for (int i = 0; i < ca->len; i++) {
+        ca->rects[i].y += ca->rects[i].h;
+    }
+
+    free(ca->cells);
+    ca->cells = new;
+}
+
+void rule_150(void *data) {
+
+    struct automaton *ca = (struct automaton *) data;
+    int total_cells = (int) pow(ca->len, ca->dimension);
+   
+    /* create new array to compute states without changing current ones */
+    int *new = malloc(sizeof(int) * total_cells);
+
+    for (int i = 1; i < ca->len-1; i++) {
+        int a, b, c;
+        a = ca->cells[i-1];
+        b = ca->cells[i];
+        c = ca->cells[i+1];
+        
+        new[i] = (c & ((a & b) | ~(a | b))) | (~c & (a ^ b));
[…]

### 2018-08-19 simulation/src/main.c

@@ -32,11 +32,17 @@ int speed = 0;
 int gens = 0;
 
 struct automaton *rule30;
+struct automaton *rule54;
 struct automaton *rule90;
 struct automaton *rule110;
+struct automaton *rule150;
+struct automaton *rule182;
+struct automaton *rule232;
+struct automaton *rule250;
+
 struct automaton *conway;
-struct automaton *brian;
 struct automaton *wires;
+struct automaton *brian;
 struct automaton *langton;
 
 struct automaton *active;
@@ -105,17 +111,36 @@ int main() {
     msg_rect.x = WIN_WIDTH - msg_rect.w;
     msg_rect.y = 0;
 
+    srand(time(NULL));
+
     /* automata initialisation */
     int *config = calloc(128*128, sizeof(int));
+    for (int i = 0; i < 128*128; i++)
+        config[i] = rand() % 2;
 
     rule30 = init_automaton(512, &rule_30, 1);
     rule30->cells = config;
+
+    rule54 = init_automaton(256, &rule_54, 1);
+    rule54->cells = config;
 
     rule90 = init_automaton(512, &rule_90, 1);
     rule90->cells = config;
 
     rule110 = init_automaton(256, &rule_110, 1);
     rule110->cells = config;
+    
+    rule150 = init_automaton(256, &rule_150, 1);
+    rule150->cells = config;
+
+    rule182 = init_automaton(256, &rule_182, 1);
+    rule182->cells = config;
+
+    rule232 = init_automaton(256, &rule_232, 1);
+    rule232->cells = config;
+
+    rule250 = init_automaton(256, &rule_250, 1);
+    rule250->cells = config;
 
     wires = init_automaton(128, &wireworld, 2);
     wires->cells = config;
@@ -136,7 +161,7 @@ int main() {
     unsigned int wait_time;
 
     /* main program execution */
-    active = langton;
+    active = rule54;;
     frame_start = SDL_GetTicks();
 
     while (running) {
@@ -260,9 +285,10 @@ void render(struct automaton *ca) {
             SDL_RenderFillRect(renderer, &ca->rects[offset]);
         }
 
-    }
+        /* only show generations for 2D automata */
+        SDL_RenderCopy(renderer, texture, NULL, &msg_rect);
 
-    SDL_RenderCopy(renderer, texture, NULL, &msg_rect);
+    }
 
     SDL_RenderPresent(renderer);
 }

### 2018-08-20 paper/paper.tex
Updates to paper. Started von Neumann's CA

@@
     \cite{WolframRule110}. We first convert $110_{10}$ to binary, and then use
-    this to fill in the table
-    as above:
+    this to fill in the table as above:
 
@@
         \item Class One: rapidly converge to a uniform state e.g. Rule 0, 232
-        \item Class Two: rapidly converge to a repetitive or stable state e.g.
+        \item Class Two: rapidly converge to a stable or oscillating state e.g.
             Rule 250
-        \item Class Three: appear to remain in a random state e.g. Rule 150, 182
+        \item Class Three: appear to evolve in a chaotic fashion e.g. Rule 30,
+            150, 182
         \item Class Four: form areas of repetitive or stable states, but also
@@
     looked only at the neighbourhood of $a_{i-1}, a_i,$ and $a_{i+1}$ when
-    applying the ruleset, two-dimensional automata have a wider choice. Two of
-    the most common types of neighbourhood to consider are the Moore
-    neighbourhood and the von Neumann neighbourhood. The former includes the 8
-    cells directly surrounding the current cell being considered, while the
-    latter only focuses on the four orthogonally ajacent cells. Ocassionally,
-    one may use the extended von Neumann neighbourhood.
+    applying the set of rules, two-dimensional automata typically have a larger
+    neighbourhood to consider. Two of the most common types of neighbourhood are
+    the Moore neighbourhood and the von Neumann neighbourhood
+    \cite{neighbourhoods}. The former includes the 8 cells directly surrounding
+    the current cell being considered, while the latter only focuses on the four
+    orthogonally ajacent cells. Ocassionally, one may use the extended von
+    Neumann neighbourhood, for which the next four closest orthogonally adjacent
+    cells are also included.
+
+    Using a Moore neighbourhood and a set of states of size $k$, we can see that
+    the total number of configurations for one particular neighbourhood is
+    $k^9$, as each of the nine cells included in the neighbourhood may be in one
+    of $k$ states. For an automaton such as Conway's Game of Life, where each
+    cell is either alive or dead and a moore neighbourhood is used, there are
+    $2^9=512$ ways to configure one of these neighbourhoods alone. The general
+    equation for the number of automata for a given neighbourhood and number of
+    possible states is $k^{k^n}$, where $k$ is the number of possible states and
+    $n$ is the size of the neighbourhood used when determining the automaton's
+    evolution between time steps (including the cell being computed); in the
+    case of a two-dimensional automaton using two states and a Moore
+    neighbourhood (such as Game of Life), there are $2^{2^9}$ possible automata
+    \cite{num_rules}.
+
+    \subsection{Game of Life}
+    At this point it makes sense to talk a bit more about Conway's Game of Life.
+    It was invented in 1970 by the British mathematician John Conway, after
+    developing an interest in them after their introduction by John von Neumann
+    in the late 1940s, who aimed to discover a hypothetical machine that was
+    capable of recreating itself. Von Neumann was able to achieve this goal
+    using a complex automaton, with 29 states and complicated rules to do so
+    \cite{self-reproducing}. Von Neumann's automaton is discussed in more depth
+    later. Conway's Game of Life was thus Conway's way of simplifying von
+    Neumann's ideas, and in fact, the Game of Life is able to simulate the Game
+    of Life itself and is in fact capable of universal computation.
 
     \subsection{Life-like cellular automata}
+
     \subsection{Cyclic cellular automata}
@@
 \section{Examples}
+    \subsection{Von Neumann's automaton}
+    \subsection{Game of Life}
+    \subsection{Langton's Ant}
+    \subsection{WireWorld}
 \section{Applications}

### 2018-08-20 simulation/include/global.h

@@ -24,7 +24,41 @@ enum states {
     CONDUCTOR,
 
     /* Langton's Ant */
-    ANT
+    ANT,
+
+    /* von Neumann's */
+    U,
+    S,
+    S0,
+    S00,
+    S000,
+    S01,
+    S1,
+    S10,
+    S11,
+
+    C00,
+    C01,
+    C10,
+    C11,
+
+    Noe,
+    Noq,
+    Eoe,
+    Eoq,
+    Soe,
+    Soq,
+    Woe,
+    Woq,
+
+    Nse,
+    Nsq,
+    Ese,
+    Esq,
+    Sse,
+    Ssq,
+    Wse,
+    Wsq
 };
 
 #endif

### 2018-08-20 simulation/src/automaton.c

@@ -509,3 +509,127 @@ void langtons_ant(void *data) {
     ca->ant->x += ca->ant->dx;
     ca->ant->y += ca->ant->dy;
 }
+
+/**
+ * John von Neumann's CA
+ *
+ * States:
+ *  Ground state:
+ *      U
+ *
+ *  Sensitised states:
+ *      S
+ *      S0
+ *      S00
+ *      S000
+ *      S01
+ *      S1
+ *      S10
+ *      S11
+ *
+ *  Confluent states:
+ *      C00
+ *      C01
+ *      C10
+ *      C11
+ *
+ *  Ordinary transmission states:
+ *      Noe
+ *      Noq
+ *      Eoe
+ *      Eoq
+ *      Soe
+ *      Soq
+ *      Woe
+ *      Woq
+ *
+ *  Special transmission states:
+ *      Nse
+ *      Nsq
+ *      Ese
+ *      Esq
+ *      Sse
+ *      Ssq
+ *      Wse
+ *      Wsq
+ *
+ * Construction Rules
+ *  U => S
+ *  S -> S0
+ *      S0 -> S00
+ *          S00 -> S000
+ *              S000 -> Eoe
+ *              S000 => Noe
+ *          S00 => Woe
+ *      S0 => S01
+ *          S01 -> Soe
+ *          S01 => Ese
+ *  S => S1
+ *      S1 -> S10
+ *          S10 -> Nse
+ *          S10 => Wse
+ *      S1 => S11
+ *          S11 -> Sse
+ *          S11 => C00
+ *
+ * Destruction Rules
+ *  special => confluent: confluent ~> U
+ *  special => ordinary: ordinary ~> U
+ *  ordinary => special: special ~> U
+ */
+
+void von_neumanns(void *data) {
+
+    struct automaton *ca = (struct automaton *) data;
+    int total_cells = (int) pow(ca->len, ca->dimension);
+
+    int *new = calloc(total_cells, sizeof(int));
+    memcpy(new, ca->cells, sizeof(int) * total_cells);
+
+    for (int i = 0; i < ca->len; i++) {
+        for (int  j = 0; j < ca->len; j++) {
+
+            int curr = i * ca->len + j;
+            int neighbourhood[4];
+
+            if (i > 0)
+                neighbourhood[0] = ca->cells[curr - ca->len];
+
+            if (j > 0)
+                neighbourhood[1] = ca->cells[curr - 1];
+
+            if (j < ca->len-1)
[…]

### 2018-08-20 simulation/src/main.c

@@ -161,7 +161,7 @@ int main() {
     unsigned int wait_time;
 
     /* main program execution */
-    active = rule54;;
+    active = rule30;
     frame_start = SDL_GetTicks();
 
     while (running) {

### 2018-08-21 simulation/include/global.h
Fixed big memory leak issue (double free)

@@ -1,8 +1,8 @@
 #ifndef _GLOBAL_H
 #define _GLOBAL_H
 
-#define WIN_WIDTH 512
-#define WIN_HEIGHT 512
+#define WIN_WIDTH 800
+#define WIN_HEIGHT 800
 
 #define FRAME_RATE 60
 

### 2018-08-21 simulation/src/automaton.c

@@ -613,16 +613,16 @@ void von_neumanns(void *data) {
             case Soq:
             case Woe:
             case Woq:
-            case Noe:
-            case Noq:
-            case Eoe:
-            case Eoq:
-            case Soe:
-            case Soq:
-            case Woe:
-            case Woq:
+            case Nse:
+            case Nsq:
+            case Ese:
+            case Esq:
+            case Sse:
+            case Ssq:
+            case Wse:
+            case Wsq:
                 if (neighbourhood[0] == Soe || neighbourhood[0] == Sse)
-                    new[curr] = 
+                    new[curr] = 1;
 
 
             }

### 2018-08-21 simulation/src/main.c

@@ -18,10 +18,6 @@ const int FRAME_INTERVAL = 1 * 1000 / FRAME_RATE;
 SDL_Window *window;
 SDL_Renderer *renderer;
 
-SDL_Surface *surface;
-SDL_Texture *texture;
-SDL_Rect msg_rect;
-
 bool running = true;
 bool paused = false;
 bool step = false;
@@ -49,6 +45,8 @@ struct automaton *active;
 
 void render(struct automaton *);
 void handle_input(void);
+
+void switch_automaton(struct automaton *);
 
 float lerp(float, float, float);
 int diag(int, int, int, int);
@@ -104,55 +102,46 @@ int main() {
         fprintf(stderr, "TTF_OpenFont: %s\n", TTF_GetError());
     }
 
-    SDL_Colour white = {255, 255, 255};
-
-    msg_rect.w = 1 * TXT_WIDTH;
-    msg_rect.h = 18;
-    msg_rect.x = WIN_WIDTH - msg_rect.w;
-    msg_rect.y = 0;
-
     srand(time(NULL));
 
     /* automata initialisation */
-    int *config = calloc(128*128, sizeof(int));
-    for (int i = 0; i < 128*128; i++)
-        config[i] = rand() % 2;
+    int *config = calloc(400*400, sizeof(int));
 
-    rule30 = init_automaton(512, &rule_30, 1);
+    rule30 = init_automaton(400, &rule_30, 1);
     rule30->cells = config;
 
-    rule54 = init_automaton(256, &rule_54, 1);
+    rule54 = init_automaton(400, &rule_54, 1);
     rule54->cells = config;
 
-    rule90 = init_automaton(512, &rule_90, 1);
+    rule90 = init_automaton(400, &rule_90, 1);
     rule90->cells = config;
 
-    rule110 = init_automaton(256, &rule_110, 1);
+    rule110 = init_automaton(400, &rule_110, 1);
     rule110->cells = config;
     
-    rule150 = init_automaton(256, &rule_150, 1);
+    rule150 = init_automaton(400, &rule_150, 1);
     rule150->cells = config;
 
-    rule182 = init_automaton(256, &rule_182, 1);
+    rule182 = init_automaton(400, &rule_182, 1);
     rule182->cells = config;
 
-    rule232 = init_automaton(256, &rule_232, 1);
+    rule232 = init_automaton(400, &rule_232, 1);
     rule232->cells = config;
 
[…]

### 2018-08-21 paper/paper.tex
Updates to paper

@@
 
-    \subsection{Game of Life}
+    \subsection{Game of Life and Life-like automata}
     At this point it makes sense to talk a bit more about Conway's Game of Life.
@@
     in the late 1940s, who aimed to discover a hypothetical machine that was
-    capable of recreating itself. Von Neumann was able to achieve this goal
-    using a complex automaton, with 29 states and complicated rules to do so
-    \cite{self-reproducing}. Von Neumann's automaton is discussed in more depth
-    later. Conway's Game of Life was thus Conway's way of simplifying von
+    capable of recreating itself \cite{GameOfLife}. Von Neumann was able to
+    achieve this goal using a complex automaton, with 29 states and complicated
+    rules to do so \cite{VonNeumannCA}. Von Neumann's automaton is discussed in
+    more depth later. Conway's Game of Life was thus his way of simplifying von
     Neumann's ideas, and in fact, the Game of Life is able to simulate the Game
-    of Life itself and is in fact capable of universal computation.
+    of Life itself and is capable of universal computation.
 
-    \subsection{Life-like cellular automata}
+    The rules for Game of Life are as follows:
+
+    \begin{itemize}
+        \item If the current cell $a_{i,j}^t$ is dead, then:
+            \begin{itemize}
+                \item if the number of alive surrounding cells is 2 or 3,
+                    then $a_{i,j}$ will become alive in the next generation
+                \item else $a_{i,j}$ remains dead
+            \end{itemize}
+        \item If the current cell $a_{i,j}^t$ is alive, then:
+            \begin{itemize}
+                \item if the number of alive surrounding cells is less than 2,
+                    then $a_{i,j}$ dies as if through isolation
+                \item if the number of alive surrounding cells is greater than
+                    3, then $a_{i,j}$ dies as if through overcrowding
+                \item else $a_{i,j}$ lives on to the next generation
+            \end{itemize}
+    \end{itemize}
+
+    There are many fascinating structures which arise out of these simple rules,
+    and can be classed as follows:
+
+    \begin{itemize}
+        \item Still Life - patterns that do not change at all from one
+            generation to the next
+        \item Oscillators - patterns which generate some sort of cycle of
+            patterns in a fixed place with a specific period
+        \item Gliders and Spaceships - patterns which cycle through some finite
+            set of patterns, but whose position is shifted each time, appeaing
+            to move
+        \item Guns - repeating patterns which produce a spaceship after a finite
+            number of generations
+        \item Puffers - moving patterns which leave a trail of stable or
+            oscillating debris at regular intervals
+        \item Rakes - moving patterns which emit spaceships at regular intervals
+            as they move
+        \item Breeders - oscillating patterns which deposit guns at regular
+            intervals. Unline guns, puffers, and rakes, who each have a linear
+            growth rate, breeders exhibit exponential growth rate
+    \end{itemize}
 

### 2018-08-22 simulation/include/automaton.h
Restructured von Neumann's

@@ -7,6 +7,51 @@
 struct automaton;
 typedef void (*simulate_fn)(void *);
 
+enum states {
+
+    /* Game of Life */
+    DEAD = 0,
+    ALIVE = 1,
+
+    /* BRIAN'S BRAIN */
+    FIRING,
+    DYING,
+    
+    /* WireWorld */
+    HEAD,
+    TAIL,
+    CONDUCTOR,
+
+    /* Langton's Ant */
+    ANT,
+
+    /* von Neumann's */
+    NORTH       = 0x0,
+    EAST        = 0x1,
+    SOUTH       = 0x2,
+    WEST        = 0x3,
+    ORDINARY    = 0x4,
+    SPECIAL     = 0x8,
+    QUIESCENT   = 0x10,
+    EXCITED     = 0x20,
+
+    GROUND,
+    S,
+    S0,
+    S00,
+    S000,
+    S01,
+    S1,
+    S10,
+    S11,
+
+    C00,
+    C01,
+    C10,
+    C11,
+
+};
+
 struct ant {
     int x, y;
     int dx, dy;

### 2018-08-22 simulation/include/global.h

@@ -5,60 +5,9 @@
 #define WIN_HEIGHT 800
 
 #define FRAME_RATE 60
-
 #define TXT_WIDTH 9
 
-enum states {
-
-    /* Game of Life */
-    DEAD = 0,
-    ALIVE = 1,
-
-    /* BRIAN'S BRAIN */
-    FIRING,
-    DYING,
-    
-    /* WireWorld */
-    HEAD,
-    TAIL,
-    CONDUCTOR,
-
-    /* Langton's Ant */
-    ANT,
-
-    /* von Neumann's */
-    U,
-    S,
-    S0,
-    S00,
-    S000,
-    S01,
-    S1,
-    S10,
-    S11,
-
-    C00,
-    C01,
-    C10,
-    C11,
-
-    Noe,
-    Noq,
-    Eoe,
-    Eoq,
-    Soe,
-    Soq,
-    Woe,
-    Woq,
-
-    Nse,
-    Nsq,
-    Ese,
-    Esq,
-    Sse,
-    Ssq,
-    Wse,
-    Wsq
-};
+#define SIZE_1D 800
+#define SIZE_2D 160
 
 #endif

### 2018-08-22 simulation/src/automaton.c

@@ -86,7 +86,6 @@ void rule_30(void *data) {
         ca->rects[i].y += ca->rects[i].h;
     }
 
-    free(ca->cells);
     ca->cells = new;
 }
 
@@ -118,7 +117,6 @@ void rule_54(void *data) {
         ca->rects[i].y += ca->rects[i].h;
     }
 
-    free(ca->cells);
     ca->cells = new;
 }
 
@@ -146,7 +144,6 @@ void rule_90(void *data) {
         ca->rects[i].y += ca->rects[i].h;
     }
 
-    free(ca->cells);
     ca->cells = new;
 }
 
@@ -178,7 +175,6 @@ void rule_110(void *data) {
         ca->rects[i].y += ca->rects[i].h;
     }
 
-    free(ca->cells);
     ca->cells = new;
 }
 
@@ -206,7 +202,6 @@ void rule_150(void *data) {
         ca->rects[i].y += ca->rects[i].h;
     }
 
-    free(ca->cells);
     ca->cells = new;
 }
 
@@ -234,7 +229,6 @@ void rule_182(void *data) {
         ca->rects[i].y += ca->rects[i].h;
     }
 
-    free(ca->cells);
     ca->cells = new;
 }
 
@@ -262,7 +256,6 @@ void rule_232(void *data) {
         ca->rects[i].y += ca->rects[i].h;
     }
 
-    free(ca->cells);
     ca->cells = new;
 }
 
@@ -285,7 +278,6 @@ void rule_250(void *data) {
         ca->rects[i].y += ca->rects[i].h;
     }
 
-    free(ca->cells);
     ca->cells = new;
 }
 
@@ -605,26 +597,26 @@ void von_neumanns(void *data) {
                 neighbourhood[3] = ca->cells[curr + ca->len];
 
             switch(ca->cells[curr]) {
-            case Noe:
-            case Noq:
-            case Eoe:
-            case Eoq:
-            case Soe:
-            case Soq:
-            case Woe:
-            case Woq:
-            case Nse:
-            case Nsq:
-            case Ese:
-            case Esq:
-            case Sse:
-            case Ssq:
-            case Wse:
-            case Wsq:
-                if (neighbourhood[0] == Soe || neighbourhood[0] == Sse)
-                    new[curr] = 1;
+            case NORTH | ORDINARY | QUIESCENT:
+            case NORTH | ORDINARY | EXCITED:
+            case NORTH | SPECIAL | QUIESCENT:
[…]

### 2018-08-22 simulation/src/main.c

@@ -18,6 +18,10 @@ const int FRAME_INTERVAL = 1 * 1000 / FRAME_RATE;
 SDL_Window *window;
 SDL_Renderer *renderer;
 
+SDL_Surface *surface;
+SDL_Texture *texture;
+SDL_Rect msg_rect;
+
 bool running = true;
 bool paused = false;
 bool step = false;
@@ -26,6 +30,9 @@ int paintbrush = ALIVE;     /* the state with which to overwrite cell */
 /* simulation speed */
 int speed = 0;
 int gens = 0;
+
+/* aesthetics */
+bool gridlines = true;
 
 struct automaton *rule30;
 struct automaton *rule54;
@@ -51,7 +58,7 @@ void switch_automaton(struct automaton *);
 float lerp(float, float, float);
 int diag(int, int, int, int);
 
-int main() {
+int main(int argc, char *argv[]) {
 
     /* SDL initialisation */
     if (SDL_Init(SDL_INIT_VIDEO | SDL_INIT_AUDIO | SDL_INIT_EVENTS) != 0) {
@@ -91,71 +98,102 @@ int main() {
         exit(EXIT_FAILURE);
     }
 
-    int ticks = 0;
-    char gen_str[12];
-    int gen_len = 1;
-    int last_gen_len = gen_len;
-
     /* text setup */
     TTF_Font *font = TTF_OpenFont("consolas.ttf", 144);
     if(!font) {
         fprintf(stderr, "TTF_OpenFont: %s\n", TTF_GetError());
     }
 
+    SDL_Colour white = {255, 255, 255};
+    msg_rect.w = 1 * TXT_WIDTH;
+    msg_rect.h = 18;
+    msg_rect.x = WIN_WIDTH - msg_rect.w;
+    msg_rect.y = 0;
+
+    char gen_str[12];
+    int gen_len = 1;
+    int last_gen_len = gen_len;
+
     srand(time(NULL));
 
     /* automata initialisation */
-    int *config = calloc(400*400, sizeof(int));
+    int *config_1d = calloc(SIZE_1D, sizeof(int));
+    config_1d[400] = 1;
 
-    rule30 = init_automaton(400, &rule_30, 1);
-    rule30->cells = config;
+    int *config = calloc(SIZE_2D*SIZE_2D, sizeof(int));
 
-    rule54 = init_automaton(400, &rule_54, 1);
-    rule54->cells = config;
[…]

### 2018-08-23 paper/paper.tex
Replicator and Diamoeba and paper updates

@@
     54 is capable of universal computation -- interacting structures form, but
-    it remains to been seen whether structures useful for computation can be
-    formed. Computational universality is discussed more in-depth later on.
+    it remains to been seen whether structures useful for computation arise.
+    Computational universality is discussed more in-depth later on.
+
+    There have been attempts to classify automata more rigorously, and Culik and
+    Yu proposed four well-defined classes for such a task; membership in such
+    classes has, however, been shown to be undecidable \cite{CulikYuMembership}.
 
@@
 
-    \subsection{Cyclic cellular automata}
+    Conway originally believed that no pattern could grow infinitely, i.e. for
+    the population to grow past any upper limit for some initial configuration
+    with finite number of alive cells, and offered fifty dollars to the first
+    person who could prove or disprove this conjecture before the end of 1970.
+    In November of that year a team from the Massachusetts Institute of
+    Technology, led by Bill Gosper, found a structure which produced gliders
+    every 30 generations, and called it the Gosper Glider Gun. Later on, smaller
+    patterns were found to produce infinite growth, one of which has been proven
+    to be minimal, starting off with only ten alive cells
+    \cite{infiniteGrowth}. More of what this means for the computational power
+    of Game of Life, as well as other automata, will be discussed under
+    Computational Universality.
+
+    A cellular automaton is Life-like (similar to Game of Life) if it meets the
+    following criteria \cite{LifeLike}:
+
+    \begin{itemize}
+        \item it is two-dimensional
+        \item each cell of the automaton may be in one of two states, alive or
+            dead (or on or off) at any point
+        \item a cell's Moore neighbourhood is used to compute the cell's state
+            in the next generation
+        \item in each time step of the automaton, the next state of the cell can
+            be expressed as a funciton of the number of adjacent cells that are
+            in the alive state and of the cell's own state
+    \end{itemize}
+
+    As with elementary cellular automata, it is helpful to refer to specific
+    rulesets in a more generalised way. Where this was using the rule's binary
+    encoding for one-dimensional automata, one such convention for
+    two-dimensional automat is using a string $Bx/Sy$, where $x$ and $y$ are a
+    sequence of digits, 0 to 8, in ascending order. The presence of a digit $d$
+    in $x$ means that a dead cell with $d$ alive neighbours will become alive in
+    the next generation, while the presence of $d$ in $y$ signifies that an
+    alive cell with $d$ alive neighbours will continue to live on to the next
+    generation. For example, Game of Life is denoted B3/S23: a dead cell is
+    \textbf{B}orn if there are exactly 3 alive neighbours surrounding it, and
+    and alive cell \textbf{S}urvives if there are two or three alive cells
+    surrounding it.
+
+    \begin{figure}[h]
+        \begin{centering}
+            \begin{tabular}{cc}
+                \includegraphics[width=2in]{replicator1.png} &
+                \includegraphics[width=2in]{replicator2.png} \\
+                %(a) first & (b) second \\[6pt]
+                \includegraphics[width=2in]{replicator3.png} &
+                \includegraphics[width=2in]{replicator4.png} \\
+                %(c) third & (d) fourth \\[6pt]
+                %\multicolumn{2}{c}{\includegraphics[width=65mm]{it} }\\
+                %\multicolumn{2}{c}{(e) fifth}
+            \end{tabular}
+            \caption{Replicator (B1357/S1357) replicating itself over the course of
+            56 generations}
+        \end{centering}
+    \end{figure}
+
+    \subsection{Other types of automata}
+    \subsubsection{Reversible cellular automata}
+    A cellular automaton is reversible if, for every configuration of the
+    automaton, exactly one configuration that causes it (its preimage). If we
+    consider the automaton as a function that maps configurations to
+    configurations, then for a reversible automaton this function would be
+    bijective. Where this is not the case, where not every configuration has a
+    preimage, these configurations without preimages are called Garden of Eden
+    patterns \cite{GardenOfEden}.
+    \subsubsection{Totalistic cellular automata}
+    \subsubsection{Proabilistic cellular automata}
+    \subsubsection{Cyclic cellular automata}
+
+\section{Computational Universality}
 
@@
     \subsection{Game of Life}
+    \subsection{Fractran}
     \subsection{Langton's Ant}
@@
 \section{Applications}
-\section{Computational Universality}
+    \subsection{Cryptography}
 \section{Firing Squad Synchronisation Problem}

### 2018-08-23 simulation/include/automaton.h

@@ -84,8 +84,10 @@ void rule_232(void *);
 void rule_250(void *);
 
 void brians_brain(void *);
+void diamoeba(void *);
 void game_of_life(void *);
-void wireworld(void *);
 void langtons_ant(void *);
+void replicator(void *);
+void wireworld(void *);
 
 #endif

### 2018-08-23 simulation/src/automaton.c

@@ -503,6 +503,163 @@ void langtons_ant(void *data) {
 }
 
 /**
+ * B1357/S1357
+ */
+void replicator(void *data) {
+
+    struct automaton *ca = (struct automaton *) data;
+    int total_cells = (int) pow(ca->len, ca->dimension);
+
+    int *new = calloc(total_cells, sizeof(int));
+    memcpy(new, ca->cells, sizeof(int) * total_cells);
+
+    for (int i = 0; i < ca->len; i++) {
+        for (int  j = 0; j < ca->len; j++) {
+
+            int curr = i * ca->len + j;
+            int alive = 0;
+
+            if (i > 0 && j > 0)
+                alive += ca->cells[curr - ca->len - 1];
+
+            if (i > 0)
+                alive += ca->cells[curr - ca->len];
+
+            if (i > 0 && j < ca->len-1)
+                alive += ca->cells[curr - ca->len + 1];
+
+            if (j > 0)
+                alive += ca->cells[curr - 1];
+
+            if (j < ca->len-1)
+                alive += ca->cells[curr + 1];
+
+            if (i < ca->len-1 && j > 0)
+                alive += ca->cells[curr + ca->len - 1];
+
+            if (i < ca->len-1)
+                alive += ca->cells[curr + ca->len];
+
+            if (i < ca->len-1 && j < ca->len-1)
+                alive += ca->cells[curr + ca->len + 1];
+
+            if (ca->cells[curr] == ALIVE) {
+
+                /* survival / death */
+                switch (alive) {
+                case 1:
+                case 3:
+                case 5:
+                case 7:
+                    new[curr] = ALIVE;
+                    break;
+                default:
+                    new[curr] = DEAD;
+                }
+
+            } else {
+
+                /* reproduction */
+                switch (alive) {
+                case 1:
+                case 3:
+                case 5:
+                case 7:
+                    new[curr] = ALIVE;
+                    break;
+                default:
+                    new[curr] = DEAD;
+                }
+
+            }
+
+        }
+    }
+
+    free(ca->cells);
+    ca->cells = new;
+}
+
+/**
+ * B35678/S5678
+ */
+void diamoeba(void *data) {
+
+    struct automaton *ca = (struct automaton *) data;
+    int total_cells = (int) pow(ca->len, ca->dimension);
+
+    int *new = calloc(total_cells, sizeof(int));
+    memcpy(new, ca->cells, sizeof(int) * total_cells);
+
+    for (int i = 0; i < ca->len; i++) {
[…]

### 2018-08-23 simulation/src/main.c

@@ -10,6 +10,7 @@
 #include "global.h"
 #include "automaton.h"
 
+#define AUTOMATA 6
 #define max(a,b) ((a) >= (b) ? a : b)
 
 /* milliseconds for each frame to take */
@@ -43,17 +44,24 @@ struct automaton *rule182;
 struct automaton *rule232;
 struct automaton *rule250;
 
-struct automaton *conway;
-struct automaton *wires;
+struct automaton *amoeboid;
 struct automaton *brian;
+struct automaton *conway;
 struct automaton *langton;
+struct automaton *replicate;
+struct automaton *wires;
+
+struct automaton *automata[6];
+int current_ca = 0;
 
 struct automaton *active;
 
+void switch_to_automaton(struct automaton *);
+int switch_to_elem(int);
+
 void render(struct automaton *);
 void handle_input(void);
-
-void switch_automaton(struct automaton *);
+void print_message(int);
 
 float lerp(float, float, float);
 int diag(int, int, int, int);
@@ -146,19 +154,32 @@ int main(int argc, char *argv[]) {
     rule250 = init_automaton(SIZE_1D, &rule_250, 1);
     rule250->cells = config_1d;
 
-    wires = init_automaton(SIZE_2D, &wireworld, 2);
-    wires->cells = config;
+    amoeboid = init_automaton(SIZE_2D, &diamoeba, 2);
+    amoeboid->cells = config;
+
+    brian = init_automaton(SIZE_2D, &brians_brain, 2);
+    brian->cells = config;
 
     conway = init_automaton(SIZE_2D, &game_of_life, 2);
     conway->cells = config;
 
-    brian = init_automaton(SIZE_2D, &brians_brain, 2);
-    brian->cells = config;
+    replicate = init_automaton(SIZE_2D, &replicator, 2);
+    replicate->cells = config;
+
+    wires = init_automaton(SIZE_2D, &wireworld, 2);
+    wires->cells = config;
 
     langton = init_automaton(SIZE_2D, &langtons_ant, 2);
     struct ant *ant = init_ant(100, 100, 0, -1, langton);
     langton->cells = config;
     langton->ant = ant;
+
+    automata[0] = conway;
+    automata[1] = wires;
+    automata[2] = brian;
+    automata[3] = langton;
+    automata[4] = amoeboid;
+    automata[5] = replicate;
 
     /* fps regulation */
     unsigned int frame_start;
@@ -176,7 +197,7 @@ int main(int argc, char *argv[]) {
[…]

### 2018-08-26 paper/paper.tex
Paper updates, added more life-like rules

@@
     automata). For example, take Rule 110, introduced by Wolfram 1983
-    \cite{WolframRule110}. We first convert $110_{10}$ to binary, and then use
-    this to fill in the table as above:
+    \cite{WolframRule110}. We first convert $110_{\text{dec}}$ to binary, and
+    then use this to fill in the table as above:
 
@@
     patterns \cite{GardenOfEden}.
+
+    There are algorithms to compute whether or not a one-dimensional cellular
+    automaton is reversible \cite{oneDReversible}; the same is not true for
+    higher dimensions however, and computing reversibility for cellular automata
+    in two-dimensions and up is undecidable \cite{twoDUndecidable}.
+
     \subsubsection{Totalistic cellular automata}
-    \subsubsection{Proabilistic cellular automata}
+    A cellular automaton is totalistic if the value of cell $a$ at time $t$
+    depends on the sum of the values of the cells in its neighbourhood at time
+    $t-1$, with the values of cells being taken from some finite set
+    \cite{totalistic}. An automaton is outer totalistic if it takes into account
+    itself as well as the total of its neighbours at time $t-1$; Conway's Game
+    of Life is an example of an outer totalistic cellular automaton where all
+    cell values are 0 or 1, as are all other Life-like automata.
+
+    \subsubsection{Probabilistic cellular automata}
     \subsubsection{Cyclic cellular automata}
@@
 \section{Computational Universality}
+    A system of rules, be they from a computer's instruction set, programming
+    language, or cellular automaton, is called computationally universal, or
+    Turing-complete, if it can be used to simulate a Turing machine.
+
+    \subsection{Turing Machines}
+    A Turing machine is a mathematical model of computation which was first
+    proposed by Alan Turing in 1936 \cite{SipserToC}. It consists of a infinite
+    tape divided into cells each of which may contain one character from some
+    finite alphabet, and a printer head capable of reading from and writing to the
+    tape on the cell at which it is currently positioned. In a more formal
+    sense, it is the seven-tuple $M=(Q, \Sigma, \Gamma, \delta, q_0,
+    q_{\text{accept}}, q_{\text{reject}})$, where $Q, \Sigma, \Gamma$ are all
+    finite sets and
+
+    \begin{itemize}
+        \item $Q$ is the set of states
+        \item $\Sigma$ is the input alphabet not containing the blank symbol,
+            $\textvisiblespace$
+        \item $\Gamma$ is the tape alphabet where $\textvisiblespace \in \Gamma$
+            and $\Sigma \subseteq \Gamma$
+        \item $\delta: Q \times \Gamma \rightarrow Q \times \Gamma \times
+            \{L,R\}$ is the transition function
+        \item $q_0 \in Q$ is the start state
+        \item $q_{\text{accept}} \in Q$ is the accept state, and
+        \item $q_{\text{reject}} \in Q$ is the reject state, with
+            $q_{\text{accept}} \neq q_{\text{reject}}$
+    \end{itemize}
 

### 2018-08-26 simulation/include/automaton.h

@@ -83,11 +83,15 @@ void rule_182(void *);
 void rule_232(void *);
 void rule_250(void *);
 
+void anneal(void *);
 void brians_brain(void *);
 void diamoeba(void *);
 void game_of_life(void *);
+void move(void *);
 void langtons_ant(void *);
+void life_wo_death(void *);
 void replicator(void *);
+void two_by_two(void *);
 void wireworld(void *);
 
 #endif

### 2018-08-26 simulation/src/automaton.c

@@ -581,6 +581,71 @@ void replicator(void *data) {
 }
 
 /**
+ * B3/S012345678
+ */
+void life_wo_death(void *data) {
+
+    struct automaton *ca = (struct automaton *) data;
+    int total_cells = (int) pow(ca->len, ca->dimension);
+
+    int *new = calloc(total_cells, sizeof(int));
+    memcpy(new, ca->cells, sizeof(int) * total_cells);
+
+    for (int i = 0; i < ca->len; i++) {
+        for (int  j = 0; j < ca->len; j++) {
+
+            int curr = i * ca->len + j;
+            int alive = 0;
+
+            if (i > 0 && j > 0)
+                alive += ca->cells[curr - ca->len - 1];
+
+            if (i > 0)
+                alive += ca->cells[curr - ca->len];
+
+            if (i > 0 && j < ca->len-1)
+                alive += ca->cells[curr - ca->len + 1];
+
+            if (j > 0)
+                alive += ca->cells[curr - 1];
+
+            if (j < ca->len-1)
+                alive += ca->cells[curr + 1];
+
+            if (i < ca->len-1 && j > 0)
+                alive += ca->cells[curr + ca->len - 1];
+
+            if (i < ca->len-1)
+                alive += ca->cells[curr + ca->len];
+
+            if (i < ca->len-1 && j < ca->len-1)
+                alive += ca->cells[curr + ca->len + 1];
+
+            if (ca->cells[curr] == ALIVE) {
+
+                    new[curr] = ALIVE;
+
+            } else {
+
+                /* reproduction */
+                switch (alive) {
+                case 3:
+                    new[curr] = ALIVE;
+                    break;
+                default:
+                    new[curr] = DEAD;
+                }
+
+            }
+
+        }
+    }
+
+    free(ca->cells);
+    ca->cells = new;
+}
+
+/**
  * B35678/S5678
  */
 void diamoeba(void *data) {
@@ -648,6 +713,237 @@ void diamoeba(void *data) {
                     break;
                 default:
                     new[curr] = DEAD;
+                }
+
+            }
+
+        }
+    }
+
+    free(ca->cells);
+    ca->cells = new;
+}
+
+/**
+ * B36/S125
+ */
+void two_by_two(void *data) {
+
+    struct automaton *ca = (struct automaton *) data;
+    int total_cells = (int) pow(ca->len, ca->dimension);
+
[…]

### 2018-08-26 simulation/src/main.c

@@ -45,10 +45,14 @@ struct automaton *rule232;
 struct automaton *rule250;
 
 struct automaton *amoeboid;
+struct automaton *annealer;
 struct automaton *brian;
 struct automaton *conway;
 struct automaton *langton;
+struct automaton *lwodeath;
+struct automaton *morley;
 struct automaton *replicate;
+struct automaton *twotwo;
 struct automaton *wires;
 
 struct automaton *automata[6];
@@ -157,14 +161,26 @@ int main(int argc, char *argv[]) {
     amoeboid = init_automaton(SIZE_2D, &diamoeba, 2);
     amoeboid->cells = config;
 
+    annealer = init_automaton(SIZE_2D, &anneal, 2);
+    annealer->cells = config;
+
     brian = init_automaton(SIZE_2D, &brians_brain, 2);
     brian->cells = config;
 
     conway = init_automaton(SIZE_2D, &game_of_life, 2);
     conway->cells = config;
 
+    lwodeath = init_automaton(SIZE_2D, &life_wo_death, 2);
+    lwodeath->cells = config;
+
+    morley = init_automaton(SIZE_2D, &move, 2);
+    morley->cells = config;
+
     replicate = init_automaton(SIZE_2D, &replicator, 2);
     replicate->cells = config;
+
+    twotwo = init_automaton(SIZE_2D, &two_by_two, 2);
+    twotwo->cells = config;
 
     wires = init_automaton(SIZE_2D, &wireworld, 2);
     wires->cells = config;
@@ -190,15 +206,7 @@ int main(int argc, char *argv[]) {
     int ticks = 0;
 
     /* main program execution */
-    if (argc == 2) {
-        if (atoi(argv[1]) == 1) {
-            active = rule30;
-        } else {
-            active = conway;
-        }
-    } else {
-        active = amoeboid;
-    }
+    active = morley;
 
     frame_start = SDL_GetTicks();
 

### 2018-08-27 simulation/src/main.c
Initialised automaton array correctly

@@ -10,7 +10,7 @@
 #include "global.h"
 #include "automaton.h"
 
-#define AUTOMATA 6
+#define AUTOMATA 10
 #define max(a,b) ((a) >= (b) ? a : b)
 
 /* milliseconds for each frame to take */
@@ -55,7 +55,7 @@ struct automaton *replicate;
 struct automaton *twotwo;
 struct automaton *wires;
 
-struct automaton *automata[6];
+struct automaton *automata[10];
 int current_ca = 0;
 
 struct automaton *active;
@@ -196,6 +196,10 @@ int main(int argc, char *argv[]) {
     automata[3] = langton;
     automata[4] = amoeboid;
     automata[5] = replicate;
+    automata[6] = morley;
+    automata[7] = lwodeath;
+    automata[8] = twotwo;
+    automata[9] = annealer;
 
     /* fps regulation */
     unsigned int frame_start;
@@ -206,7 +210,7 @@ int main(int argc, char *argv[]) {
     int ticks = 0;
 
     /* main program execution */
-    active = morley;
+    active = automata[0];
 
     frame_start = SDL_GetTicks();
 
@@ -655,6 +659,18 @@ void print_message(int n) {
         break;
     case 5:
         strcat(msg, "Replicator");
+        break;
+    case 6:
+        strcat(msg, "Morley's automaton");
+        break;
+    case 7:
+        strcat(msg, "Life without Death");
+        break;
+    case 8:
+        strcat(msg, "Two By Two");
+        break;
+    case 9:
+        strcat(msg, "Anneal");
         break;
     }
     strcat(msg, "!\n");

### 2018-09-13 paper/paper.tex
Additions to paper

@@
 \usepackage{diagbox}
+\usepackage{enumitem}
+\usepackage{braket}
+
+\usepackage{tikz}
+\usetikzlibrary{chains} % TM tape
+\usetikzlibrary{automata,positioning}   % automaton diagram
+
+\newenvironment{turing}[2]{
+    \begin{enumerate}[leftmargin=0pt,labelsep=0pt,align=left,parsep=0pt]
+        \item[$#1={}$]``\ignorespaces#2
+            \begin{enumerate}[
+                nosep,
+                align=left,
+                labelwidth=1.5em,
+                label=\bfseries\arabic{*}.,
+                ref=\arabic{*}
+            ]}{\unskip''
+            \end{enumerate}
+    \end{enumerate}}
+
+\newcommand{\bitem}{\item\hspace*{1em}\ignorespaces}
 
@@
     chaotic background. This starting chaos is modelled in the cellular
-    automaton by randomly assigning each cell a value of 0 or 1.
+    automaton by randomly assigning each cell a value of 0 or 1, with each
+    occurring with probability one half.
 
@@
     Reading the bottom row of this transition table, we get the binary number
-    $01011010_{\text{bin}} = 90_{\text{dec}}$, allowing us to name this rule set
-    Rule 90. Using this convention, we may now construct our own automata based
+    $01011010_{\text{bin}} = 90_{\text{dec}}$, and so we name this rule set Rule
+    90. Using this convention, we may now construct our own automata based
     solely on the knowledge of its name (at least for elementary cellular
@@
     indeed many automata are equivalent to each other thanks to simple
-    transformations of theirunerlying geometry. The first of these
+    transformations of their underlying geometry. The first of these
     transformations is reflection in the vertical axis; the result of applying
@@
     below gives that of our new ruleset, made by applying the mirror rule to
-    Rule 30. Thus we get Rule 30's mirror rule Rule $01010110_{\text{bin}}$ =
+    Rule 30. Thus we get Rule 30's mirror rule, Rule $01010110_{\text{bin}}$ =
     Rule $86_{\text{dec}}$. Rules which generate the same ruleset under
@@
     the current cell being considered, while the latter only focuses on the four
-    orthogonally ajacent cells. Ocassionally, one may use the extended von
+    orthogonally ajacent cells. Occasionally, one may use the extended von
     Neumann neighbourhood, for which the next four closest orthogonally adjacent
@@
     rules to do so \cite{VonNeumannCA}. Von Neumann's automaton is discussed in
-    more depth later. Conway's Game of Life was thus his way of simplifying von
-    Neumann's ideas, and in fact, the Game of Life is able to simulate the Game
-    of Life itself and is capable of universal computation.
-
-    The rules for Game of Life are as follows:
+    more depth later. The Game of Life was thus Conway's way of simplifying von
+    Neumann's ideas; in fact, not only can the Game of Life simulate the Game of
+    Life, it is also capable of universal computation, a topic discussed more
+    in-depth later. The rules for Game of Life are as follows:
 
@@
     \end{itemize}
+
+    On 6\textsuperscript{th} March, 2018 the first knightship was discovered by
+    Adam P. Goucher, named Sir Robin \cite{Knightship}. A knightship is a
+    spaceship which moves two squares horizontally for every square it moves
+    vertically, as opposed to regular spaceships, which move only vertically or
+    horizontally, or gliders, which move exactly diagonally. Sir Robin became
+    the first new spaceship movement pattern for an elementary spaceship to be
+    discovered in forty eight years \cite{fortyeight}.
 
@@
                 \includegraphics[width=2in]{replicator2.png} \\
-                %(a) first & (b) second \\[6pt]
                 \includegraphics[width=2in]{replicator3.png} &
                 \includegraphics[width=2in]{replicator4.png} \\
-                %(c) third & (d) fourth \\[6pt]
-                %\multicolumn{2}{c}{\includegraphics[width=65mm]{it} }\\
-                %\multicolumn{2}{c}{(e) fifth}
             \end{tabular}
@@
     language, or cellular automaton, is called computationally universal, or
-    Turing-complete, if it can be used to simulate a Turing machine.
+    Turing-complete, if it can be used to simulate a Turing machine. Before
+    getting into the computational universality of cellular automata, we need to
+    discuss the concept of Turing machines and what it means for a function to
+    be computable.
 
@@
         \item $\Sigma$ is the input alphabet not containing the blank symbol,
-            $\textvisiblespace$
-        \item $\Gamma$ is the tape alphabet where $\textvisiblespace \in \Gamma$
+            $\sqcup$
+        \item $\Gamma$ is the tape alphabet, where $\sqcup \in \Gamma$
             and $\Sigma \subseteq \Gamma$
         \item $\delta: Q \times \Gamma \rightarrow Q \times \Gamma \times
-            \{L,R\}$ is the transition function
+            \{L,R,S\}$ is the transition function, signifying whether the tape
+            head should move left, right, or stay in place, respectively
         \item $q_0 \in Q$ is the start state
@@
     \end{itemize}
+
+    It is conventional to use $\vdash$ to signify the start of the input string,
+    and an infinite string of $\sqcup$ to show the end of the input string. For
+    example, the string 01011010 we would expect to be represented on the tape
+    as:
+
+    \begin{center}
+        \begin{tikzpicture}
+            \edef\sizetape{0.7cm}
+            \tikzstyle{tmtape}=[draw,minimum size=\sizetape]
+            \tikzstyle{tmhead}=[arrow box,draw,minimum size=.5cm,arrow box
+            arrows={east:.25cm, west:0.25cm}]
+
+            \begin{scope}[start chain=1 going right,node distance=-0.15mm]
+                \node [on chain=1,tmtape,draw=none] {$\ldots$};
+                \node [on chain=1,tmtape] {$\vdash$};
+                \node [on chain=1,tmtape] {0};
+                \node [on chain=1,tmtape] {1};
+                \node [on chain=1,tmtape] {0};
+                \node [on chain=1,tmtape] {1};
+                \node [on chain=1,tmtape] {1};
+                \node [on chain=1,tmtape] {0};
+                \node [on chain=1,tmtape] {1};
+                \node [on chain=1,tmtape] {0};
+                \node [on chain=1,tmtape] {$\sqcup$};
+                \node [on chain=1,tmtape] {$\sqcup$};
+                \node [on chain=1,tmtape] {$\sqcup$};
+                \node [on chain=1,tmtape,draw=none] {$\ldots$};
+            \end{scope}
+        \end{tikzpicture}
+    \end{center}
+
+    We call a set of strings Turing-recognisable \cite{Recognisable} if we can
+    construct some Turing machine to recognise it, that is, every input we feed
+    it that is in the language will cause the Turing machine to eventually move
+    to the accept state. Notice, however, that while there are two
+    ``destination'' states, $q_{\text{accept}}$ and $q_{\text{reject}}$, there
+    is a third possible outcome: that our machine does not halt, but instead
+    loops somewhere in its proceedings infinitely. Thus we call a set of strings
+    decidable if our Turing machine always halts for its inputs, that is, for
+    any input we feed it we are guaranteed to land in either $q_{\text{accept}}$
+    or $q_{\text{reject}}$.
+
+    As an example of a Turing machine in action, suppose we want to construct a
+    one to accept the language $L = \{w\#w | w \in \{0,1\}\ast\}$.  This is not
+    Context Free \cite{nonCFL}, hinting at the power behind Turing Machines with
+    regard to computability. We construct the machine informally as follows:
+
+    \begin{turing}{M}{on input $w$:}
+        \item Find a decomposition of $w$ into $w_1 \# w_2$
+        \item Start at the first symbol of $w_1$, replace the symbol with a
+            cross, and skip forward to the first symbol of $w_2$.
+            \bitem if this symbol is the same as the one you just crossed off,
+            cross it off and rewind to the symbol on the right of the first
+            symbol of $w_1$
+            \bitem else reject
+        \item Repeat this process, each time skipping forward and rewinding so
+            that you are inspecting the first symbol to the right of a cross
+            each time
+        \item If you end up inspecting a cell with the symbol $\#$, and the
+            symbols either side of it are crosses, accept
+    \end{turing}
+
+    The final thing to cover in order to understand how cellular automata such
+    as the Game of Life can be used to simulate a Turing Machine is the concept
+    of a Universal Turing Machine. A Universal Turing Machine is one which
+    recognises the language
+
+    $$A_{TM} = \{\braket{M,w} | M \text{ is a Turing Machine and } M
+    \text{ accepts } w\} $$
+
+    The $\braket{M,w}$ notation is used to denote the encoding of the Turing
+    Machine $M$ and its input $w$ in a way that some Turing Machine would be
+    able to understand and use as input. In essence, Universal Turing Machines
+    are those which are capable of simulating any other Turing Machine just from
+    its description and an input, and it is from this angle that we will look at
+    the universality of certain cellular automata.
+
+    \begin{figure}
+        \centering
+        \begin{tikzpicture}[
+                ->,
+                node distance=3cm,
+                on grid,
+                auto]
+            \node[state,initial]    (q1) []                     {$q_1$};
+            \node[state]            (q2) [above right of=q1]    {$q_2$};
+            \node[state]            (q3) [right of=q2]          {$q_3$};
+            \node[state]            (q4) [below right of=q1]    {$q_4$};
+            \node[state]            (q5) [right of=q4]          {$q_5$};
+            \node[state]            (q6) [above right of=q5]    {$q_6$};
+            \node[state]            (q7) [right of=q6]          {$q_7$};
+            \node[state,accepting]  (qa) [right of=q7]          {$q_a$};
+            \node[state,accepting]  (qr1) [right of=q3]         {$q_r$};
+            \node[state,accepting]  (qr2) [right of=q5]         {$q_r$};
+
+            \path[->]
+            (q1)
+            edge [loop above] node [align=center] {\tiny$\vdash \rightarrow R$
+            \\ \tiny$\times \rightarrow R$} ()
+            edge  node  [swap] {\tiny$0 \rightarrow \times, R$} (q2)
+            edge  node  {\tiny$1 \rightarrow \times, R$} (q4)
+            (q2)
+            edge [loop above] node [align=center] {\tiny$0 \rightarrow R$ \\
+            \tiny$1 \rightarrow R$} ()
+            edge node {\tiny$\# \rightarrow R$} (q3)
+            (q3)
+            edge [loop above] node {\tiny$\times \rightarrow R$} ()
+            edge node [swap] {\tiny$0 \rightarrow \times, L$} (q6)
+            edge node {\tiny$1 \rightarrow \times, S$} (qr1)
+            (q4)
+            edge [loop below] node [align=center] {\tiny$0 \rightarrow R$ \\
+            \tiny$1 \rightarrow R$} ()
+            edge node {\tiny$\# \rightarrow R$} (q5)
+            (q5)
+            edge [loop below] node {\tiny$\times \rightarrow R$} ()
+            edge node {\tiny$0 \rightarrow \times, S$} (qr2)
+            edge node {\tiny$1 \rightarrow \times, L$} (q6)
+            (q6)
+            edge [loop above] node [align=center] {\tiny$0 \rightarrow L$ \\
+            \tiny$1 \rightarrow L$} ()
+            edge [loop below] node [align=center] {\tiny$\times \rightarrow L$} ()
+            edge [bend left] node {\tiny$\# \rightarrow L$} (q7)
+            edge node [swap] {\tiny$\vdash \rightarrow R$} (q1)
+            (q7)
+            edge [bend left] node [align=center] {\tiny$0 \rightarrow L$ \\
+            \tiny$1 \rightarrow L$} (q6)
+            edge node {\tiny$\times \rightarrow S$} (qa);
+        \end{tikzpicture}
+        \caption{A possible Turing Machine to recognise $L$}
+        \label{TM:L}
+    \end{figure}
+
+    \subsection{Computability and Universality of Cellular Automata}
+    A Turing Machine may compute a function by starting with the input to the
+    function on the tape and halting with the output of the function on the
+    tape. A function $f : \Sigma \ast \rightarrow \Sigma \ast$ is computable if
+    some Turing Machine $M$ halts, for every input $w$, with just the result
+    $f(w)$ on the tape \cite{Computable}. Thus, any model we can devise which
+    could simulate the run of a Turing Machine on some input would be capable of
+    computing all the functions that the Turing Machine could, and hence such a
+    model would be Turing complete.
 

### 2018-09-13 paper/paper.tex
Finished intro to universality

@@
     one to accept the language $L = \{w\#w | w \in \{0,1\}\ast\}$.  This is not
-    Context Free \cite{nonCFL}, hinting at the power behind Turing Machines with
-    regard to computability. We construct the machine informally as follows:
+    Context Free \cite{nonCFL}, meaning Turing Machines are at least more
+    powerful than Push Down Automata. We construct the machine informally as
+    follows:
 
@@
     its description and an input, and it is from this angle that we will look at
-    the universality of certain cellular automata.
+    the universality of certain cellular automata. In case this seems like a bit
+    of a leap, we can simply describe the Universal Turing Machine $U$ as
+    follows:
+
+    \begin{turing}{U}{on input $\braket{M,w}$, where $M$ is a Turing Machine and
+        $w$ is a string:}
+        \item Run $M$ on input $w$
+        \item If $M$ ever enters its accept state, accept; if $M$ ever enters
+            its reject state, reject
+    \end{turing}
+
+    Note that $U$ will never halt if $M$ never halts, and so while it recognises
+    the language $A_{TM}$, it does not decide it. In fact, $A_{TM}$ is called
+    the Halting Problem and is undecidable: there is no way for an algorithm to
+    determine whether some Turing Machine $M$ may or may not halt on input $w$,
+    without simulating it itself.
 
@@
             (q1)
-            edge [loop above] node [align=center] {\tiny$\vdash \rightarrow R$
-            \\ \tiny$\times \rightarrow R$} ()
+                edge [loop above] node [align=center] {\tiny$\vdash
+                    \rightarrow R$ \\ \tiny$\times \rightarrow R$} ()
                 edge  node  [swap] {\tiny$0 \rightarrow \times, R$} (q2)
@@
 
-    \subsection{Computability and Universality of Cellular Automata}
+    \subsection{Universality of Game of Life}
     A Turing Machine may compute a function by starting with the input to the
@@
 
-\section{Examples}
+    The key to Game of Life's universal computation is its ability to create
+    infinite growth, for example by using glider guns. Guns can then be
+    positioned to shoot at blocks in specific locations, and doing so correctly
+    can be used to move the block closer or further away. This sliding block
+    memory can be used to simulate a counter \cite{Counter}. Logical AND, OR,
+    and NOT gates can also be constructed using gliders, with the presence of a
+    glider signifying a 1, and 0 otherwise. These three Boolean operators are
+    theoretically necessary and sufficient to build a Universal Turing Machine
+    \cite{BoolTM}. A fully functioning Turing Machine was designed to completion
+    by Paul Rendell on 2\textsuperscript{nd} April, 2000 \cite{golTM}, which
+    doubled the number on the input tape. Further extensions led to the
+    development of Universal Turing Machines in February 2010 and March 2011,
+    both of which were capable of simulating the Turing Machine from 2000
+    \cite{golUTM1,golUTM2}.
+
+    \begin{figure}[h]
+        \includegraphics{}
+        \caption{}
+        \label{}
+    \end{figure}
+
+\section{More cellular automata}
+    \subsection{WireWorld}
+    \subsubsection{Computational Universality}
+    \subsection{Rule 110}
+    \subsubsection{Computational Universality}
     \subsection{Von Neumann's automaton}
-    \subsection{Game of Life}
     \subsection{Fractran}
     \subsection{Langton's Ant}
-    \subsection{WireWorld}
 \section{Applications}

### 2018-09-18 paper/paper.tex
Oops last was old version
Paper updoots

@@
 \usepackage[margin=1in]{geometry}
+
 \usepackage{graphicx}
+\graphicspath{{figs/}}
+
 \usepackage{subfig}
@@
 
-    \subsection{Universality of Game of Life}
+    \subsection{Universality in the Conway's Game of Life}
     A Turing Machine may compute a function by starting with the input to the
@@
     some Turing Machine $M$ halts, for every input $w$, with just the result
-    $f(w)$ on the tape \cite{Computable}. Thus, any model we can devise which
-    could simulate the run of a Turing Machine on some input would be capable of
-    computing all the functions that the Turing Machine could, and hence such a
-    model would be Turing complete.
+    $f(w)$ on the tape \cite{Computable}. So for each computable function, there
+    is a Turing Machine $T$ such that running $T$ on input $x$ always halts and
+    outputs the correct answer (i.e., ends with the correct answer on the tape).
+    Thus, any model we can devise which could simulate the run of a Turing
+    Machine on some input would be capable of computing all the functions that
+    the Turing Machine could, that is, all computable functions, and hence such
+    a model would be Turing complete.
+
+    Processor instruction sets may be fully implemented using just a NOT gate
+    and one of either AND or OR in order to compute all binary logic. All that
+    remains to obtain Turing Completeness is access to infinite, random access
+    memory. Of course, given the finite nature of computers, this is impossible,
+    but as long as the memory is as large as is required by the Turing Machine,
+    this will suffice. Memory can be implemented in logic circuits using D-type
+    flip flops, which can be done with four NAND gates. Thus, if this can be
+    implemented using a cellular automaton, that automaton will be a universal
+    computer.
 
@@
     \begin{figure}[h]
-        \includegraphics{}
-        \caption{}
-        \label{}
+        \centering
+        \includegraphics[width=5in]{golTM.png}
+        \caption{Section of a Turing Machine implemented in Conway's Game of
+        Life}
+        \label{fig:golTM}
     \end{figure}
@@
     \subsection{WireWorld}
+    WireWorld is a two-dimensional cellular automaton first proposed by Brian
+    Silverman in 1987. The ruleset is relatively simpleset, with four distinct
+    states:
+    \begin{enumerate}
+        \item empty
+        \item conductor
+        \item electron head
+        \item electron tail
+    \end{enumerate}
+
+    As the state names may betray, WireWorld is particularly good at simulating
+    electronics. Like Game of Life, WireWorld uses a Moore neighbourhoodin its
+    calculations, as opposed to the more restricted Von Neumann
+    neighbourhood.The rules are as follows:
+    \begin{itemize}
+        \item $empty \rightarrow empty$
+        \item $head \rightarrow tail$
+        \item $tail \rightarrow conductor$
+        \item $conductor \rightarrow head \text{ if exactly 2 neighbours are
+            $head$, else } conductor$
+    \end{itemize}
+
+    \begin{figure}[h]%
+        \centering \subfloat[OR and XOR gates]{{\includegraphics[width=2in]{wireworld_gates.png}}}%
+        \qquad
+        \subfloat[D-type flip flop]{{\includegraphics[width=2in]{wireworld_flipflop.png}}}%
+        \caption{All the structures required for universal computation implemented in WireWorld}%
+        \label{fig:nature_fractals}%
+    \end{figure}
+
     \subsubsection{Computational Universality}
+    As with Game of Life, WireWorld can be used to implement the logic gates
+    required for Turing Completeness, and is inherently more suited to doing so.
+    A full computer has been constructed in WireWorld, which calculates prime
+    numbers \cite{WW_TM}. Langton's Ant, a cellular automaton that will be
+    discussed shortly, has also been implemented in WireWorld \cite{WW_LA}.
+
     \subsection{Rule 110}
     \subsubsection{Computational Universality}
+    The key to Rule 110's computational universality is its ability to simulate
+    a ``tag system'', which is universal, using gliders inherent to the
+    automaton \cite{Rule110_universal}. The details are somewhat beyond the
+    scope and level of this report, but Cook first proves that a tag system that
+    removes two symbols at each stage is universal by compiling a two-state
+    Turing Machine program. After this, he shows that such a tag system may be
+    implemented using these gliders in Rule 110, and consequently shows that
+    this cellular automaton is capable of universal computation. Some such
+    gliders used by Cook in his proof of Rule 110's universality are shown in
+    Fig. \ref{fig:110_gliders}.
+
+    \begin{figure}[h]
+        \centering
+        \includegraphics[width=4in]{rule110_gliders.png}
+        \caption{Gliders that interact by a) translation and b) forming a new
+        structure}
+        \label{fig:110_gliders}
+    \end{figure}
+
+    Due to Rule 110's simplicity, it suggests naturally occurring physical
+    systems may also be capable of universality (systems similar to the way the
+    shapes formed on the cone snail's shell, for example). Therefore, many of
+    the properties of such systems would be undecidable, for which it is
+    impossible to find closed-form mathematical solutions.
+
     \subsection{Von Neumann's automaton}
+    Cellular automata can trace their roots back to the work done in the late
+    1940s by Stanislaw Ulam and John von Neumann. Von Neumann wanted to
+    ultimately answer the question of whether a machine was capable of
+    replicating itself. We are used to automata creating structures much simpler
+    than themselves, for example a factory that produces bolts: presumably, one
+    section of the machine would cast the metal, while another would make the
+    thread. He was concerned whether there was some form of ``extra-mechanical
+    magic'' to self-reproduction \cite{CA_origins}.
+
+    By 1952, von Neumann had completed his goal of engineering an automaton
+    capable of recreating itself. It used a total of 29 states, and each cell's
+    state in the next generation is calculated by considering its von Neumann
+    neighbourhood in the current one. A summary of the states is as follows:
+
+    \begin{enumerate}
+        \item one ground state, $U$
+        \item eight transition states,
+            $S,S_0,S_{00},S_{000},S_{01},S_1,S_{10},S_{11}$
+        \item four confluent states, $C_{00}, C_{01}, C_{10}, C{11}$
+        \item eight ordinary transmission states in directions North, East,
+            South, and West
+            \begin{itemize}
+                \item North-quiescent, North-excited
+                \item East-quiescent, East-excited
+                \item South-quiescent, South-excited
+                \item West-quiescent, West-excited
+            \end{itemize}
+        \item eight special transmission states in the same directions
+            \begin{itemize}
+                \item North-quiescent, North-excited
+                \item East-quiescent, East-excited
+                \item South-quiescent, South-excited
+                \item West-quiescent, West-excited
+            \end{itemize}
+    \end{enumerate}
+
+    The ruleset can be broken down into four sections, concerning those for
+    construction, destruction, transmission states, and confluent states. They
+    are extensive and a full summary can be found at \cite{VN_rules}.
+
     \subsection{Fractran}
@@
 \section{Firing Squad Synchronisation Problem}
+    \subsection{The Problem}
+    Imagine a group of soldiers all part of a firing squad, with a general at
+    one end. Firing squads would aim to have all of its soldiers fire at the
+    same time so that no one soldier could be sure he had delivered the fatal
+    wound to the condemned. The soldiers of the firing squad can be modelled as
+    cells in a cellular automaton, and so the Firing Squad Synchronisation
+    Problem, or FSSP, is concerned with ensuring that all states end up in a
+    ``firing'' state at some point, at the same point in time. At each time
+    step, a soldier may change what he is doing based on what him and his two
+    immediate neighbours are doing (so the two soldiers on the ends only take
+    into account their own current action and that of his single neighbour).
+    Assume all soldiers start off in the same inactive state, also called the
+    quiescent state. The question begins when the general initially gives the
+    command to fire: assuming there is a fixed number of states, regardless of
+    how many soldiers are in the line, how can the states be defined such that
+    all soldiers are in the firing state at the same time?
+
 \section{Majority Problem}

### 2018-09-19 paper/paper.tex
Almost done

@@
     \subsection{Cryptography}
+
 \section{Firing Squad Synchronisation Problem}
     \subsection{The Problem}
-    Imagine a group of soldiers all part of a firing squad, with a general at
-    one end. Firing squads would aim to have all of its soldiers fire at the
-    same time so that no one soldier could be sure he had delivered the fatal
-    wound to the condemned. The soldiers of the firing squad can be modelled as
-    cells in a cellular automaton, and so the Firing Squad Synchronisation
-    Problem, or FSSP, is concerned with ensuring that all states end up in a
-    ``firing'' state at some point, at the same point in time. At each time
-    step, a soldier may change what he is doing based on what him and his two
-    immediate neighbours are doing (so the two soldiers on the ends only take
-    into account their own current action and that of his single neighbour).
-    Assume all soldiers start off in the same inactive state, also called the
-    quiescent state. The question begins when the general initially gives the
-    command to fire: assuming there is a fixed number of states, regardless of
-    how many soldiers are in the line, how can the states be defined such that
-    all soldiers are in the firing state at the same time?
+    A firing squad consists of a group of soldiers, all of whom will fire a shot
+    at the condemned party in order to end his or her life, at the command of a
+    general. An important factor to this is that all the soldiers fire
+    simultaneously; this means that no one soldier can be sure that they are the
+    one to have administered the fatal shot, and can thus rest easy at night. If
+    we model the line of soldiers as a one-dimensional cellular automaton, we
+    want all of the cells in this automaton to be in a ``firing'' state
+    simultaneously.  More specifically, at each time step each soldier may
+    change what he is doing (i.e. change his state) based solely upon what he
+    and his two immediate neighbours are doing. Note that the soldiers at either
+    end of the line only consider the actions of himself and his sole neighbour.
+    Furthermore, we assume that each soldier starts in the same inactive state
+    (also known as the quiescent state), and that the general is located at one
+    end of the line, and only he may give the initial command to fire. He
+    himself must also fire his weapon at the to-be recipient of the discharge of
+    such weapons. Thus the question asked by the Firing Squad Synchronisation
+    Problem is this: assuming there is a fixed number of states, and regardless
+    of how many soldiers form the firing squad, how can the states and
+    transitions be defined such that all soldiers are all, at some moment in
+    time, firing together?
+
+    \subsection{The Solution}
+    Since the signal can only travel at most one cell to the left/right between
+    each time step, it is clear that a minimal-time solution can not take shorter
+    than $2(n-1)$ time (the signal must travel all the way down the line and all
+    the way back, on the way down skipping the general and on the way back
+    bouncing straight off the end soldier). The first minimal-time solution to
+    the problem was Abraham Waksman's, published in 1966 \cite{Waksman}. This
+    solution uses the following six basic states:
+
+    \begin{enumerate}
+        \item $Q$ - the starting, quiescent state
+        \item $T$ - the final, firing state
+        \item $R$ - the trigger signal for the $B$ state:
+            \begin{itemize}
+                \item $R_0$ - propagates left
+                \item $R_1$ - propagates right
+            \end{itemize}
+        \item $B$ - the state that generates the $P$ state
+            \begin{itemize}
+                \item $B_0$ - blocks the $R$ state
+                \item $B_1$ - passes through the $R$
+            \end{itemize}
+        \item $P$ - the state generating the $A$ state (leads to $T$ if both
+            neighbours in $P$ state)
+            \begin{itemize}
+                \item $P_0$ - generates $A_{0xx}$
+                \item $P_1$ - generates $A_{1xx}$
+            \end{itemize}
+        \item $A$ - the propagating state
+            \begin{itemize}
+                \item $A_{000}, A_{001}, A_{010}, A_{011}, A_{100}, A_{101},
+                    A_{110}, A_{111}$
+                \item $A_{0xx}$ - generates the state $R$ with no delay
+                \item $A_{1xx}$ - generates the state $R$ with one unit time
+                    delay
+                \item $A_{x00}, A_{x01}$ - propagates to the left
+                \item $A_{x10}, A_{x11}$ - propagates to the right
+            \end{itemize}
+    \end{enumerate}
+
+    There are also two extra placeholder states:
+    \begin{itemize}
+        \item $\phi$ - external state, no neighbour to this side
+        \item $\gamma$ - neighbours note explicitly mentioned
+    \end{itemize}
+
+    The way Waksman's solution works can be summarized by the interactions
+    between the states of his automaton which occur:
+    \begin{itemize}
+        \item Every $A$ signal can be assigned a parity, even or odd, based on
+            whether it is at an even or odd cell. The parity is relative to the
+            cell who originated the signal
+        \item Every cell with an even $A$ signal generates a new $R$ signal that
+            travels in the opposite direction of the cell's $A$ signal
+        \item Each type of $B$ signal will switch to the other type upon
+            intersection with an $R$ signal. One type allows the $R$ signal to
+            continue propagating, the other does not
+        \item When $A$ and $B$ signals intersect, they set up a new generator,
+            or $P$ signal
+    \end{itemize}
+
+    Of course, time complexity is not the only metric by which to judge a
+    solution's merit. The first solution to the problem, proposed by John
+    McCarthy and Marvin Minsky, was relatively much more simple, involving
+    propagating two waves down the line of soldiers, one moving at three times
+    the speed of the other. The faster wave bounces off the edges and meet again
+    in the center, creating two more waves for a total of four waves. This
+    process repeats, until each cell is part of a wave simultaneously. At this
+    point, the soldiers fire. This requires $3n$ units of time for $n$ soldiers,
+    and requires 15 states.
+
+    The current best solution to the problem was devised by Jacques Mazoyer
+    \cite{FSSP_best}, and uses six states to solve the problem in time $2(n-1)$.
+
+    In minimal-time solutions, the general sends signals $S_1, S_2, S_3, ..., S_i$ to
+    the right (assuming he is the left most soldier) at speeds $1, \frac{1}{3},
+    \frac{1}{7}, ..., \frac{1}{2^{i-1}-1}$. $S_1$ reflects and meets signal
+    $S_i$ (for $i \ge 2$) at cell $\frac{n}{2^{i-1}}$. When $S_1$ reflects it also
+    creates another general at the right end, and the process repeats. Fig.
+    \ref{fig:mazoyer} shows Mazoyer's minimal-time and so far minimal-state
+    solution to FSSP.
+
+    \begin{figure}[h]
+        \centering
+        \includegraphics[width=4in]{mazoyer_fssp.png}
+        \caption{Mazoyer's minimal solution to FSSP. Time increases from left to
+        right.}
+        \label{fig:mazoyer}
+    \end{figure}
 
 \section{Majority Problem}
+    \subsection{The Problem}
+    Suppose we have a one-dimensional cellular automaton, each of whose cells
+    may be in one of two states. There are $i+j$ cells in total, $i$ of which
+    are in the zero state, the other $j$ of which are in the one state. A
+    correct solution to this problem must eventually set all cells to zero if $i
+    > j$, or one if $i < j$. The desired final state of the automaton is
+    unspecified if $i = j$.
+
+    \subsection{The Solution}
+    G\'acs, Kurdyumov, and Levin designed an automaton the solves the Majority
+    Problem in most cases, but not all. In fact, with a random starting
+    configuration, their solution is about 78\% successful in correctly
+    determining the majority \cite{GKL_majority}. They approached it by
+    classifying the quality of a particular cellular automaton by the proportion
+    of the $2^{i+j}$ possible starting configurations in which it correctly
+    identified the majority.
+
+    An overview of their method is as follows: if a cell is 0, the next state is
+    the majority value between itself, its neighbour immediately to its left,
+    and its neighbour three spaces to the left. Similarly, if the cell is a 1,
+    the value is taken as the majority of itself, its immediate right neighbour,
+    and its neighbour three spaces to the right.
+
+    Capcarrere, Sipper, and Tomassini observed \cite{CST_majority} the Majority
+    Problem may be solved perfectly by altering the definition of having
+    recognised the majority. For example, if Rule 184 is run on a finite
+    universe with cyclic boundaries, each cell will eventually see two
+    consecutive states of the majority value infinitely often, but will see two
+    consecutive states of the minority value only finitely many times. Thus the
+    Majority Problem cannot be solved perfectly if we require all cells to
+    eventually stabilise to the majority state, but can if we relax this
+    requirement and allow Rule 184 to run infinitely \cite{LB_majority}.
 
 \bibliography{bibliography}
-
 \end{document}

### 2018-09-21 paper/paper.tex
Poster finished, elementary automata improved

@@
     them the ability to simulate digital processes and the potential to solve
-    problems in this area.
+    a range of problems in this area.
 

### 2018-09-21 simulation/include/automaton.h

@@ -73,6 +73,10 @@ struct automaton {
 struct automaton *init_automaton(int len, simulate_fn func, int d);
 struct ant *init_ant(int x, int y, int dx, int dy, struct automaton *);
 
+void next_layer(struct automaton *);
+void reset_rects(struct automaton *);
+void reset_automaton(struct automaton *);
+
 /* elementary cellular automata */
 void rule_30(void *);
 void rule_54(void *);

### 2018-09-21 simulation/include/global.h

@@ -7,7 +7,7 @@
 #define FRAME_RATE 60
 #define TXT_WIDTH 9
 
-#define SIZE_1D 800
+#define SIZE_1D 400
 #define SIZE_2D 160
 
 #endif

### 2018-09-21 simulation/src/automaton.c

@@ -82,9 +82,7 @@ void rule_30(void *data) {
     new[0] = ca->cells[0] | ca->cells[1];
     new[ca->len-1] = ca->cells[ca->len-2] ^ ca->cells[ca->len-1];
 
-    for (int i = 0; i < ca->len; i++) {
-        ca->rects[i].y += ca->rects[i].h;
-    }
+    next_layer(ca);
 
     ca->cells = new;
 }
@@ -113,13 +111,35 @@ void rule_54(void *data) {
     new[0] = ca->cells[0] ^ ca->cells[1];
     new[ca->len-1] = ca->cells[ca->len-2] ^ ca->cells[ca->len-1];
 
-    for (int i = 0; i < ca->len; i++) {
-        ca->rects[i].y += ca->rects[i].h;
-    }
+    next_layer(ca);
 
     ca->cells = new;
 }
 
+void next_layer(struct automaton *ca) {
+    for (int i = 0; i < ca->len; i++) {
+        ca->rects[i].y += ca->rects[i].h;
+    }
+}
+
+void reset_rects(struct automaton *ca) {
+    for (int i = 0; i < ca->len; i++) {
+        ca->rects[i].y = 0;
+    }
+}
+
+void reset_automaton(struct automaton *ca) {
+    int arr_size = pow(ca->len, ca->dimension) * sizeof(int);
+    printf("Clearing %d bytes\n", arr_size);
+    memset(
+        ca->cells,
+        DEAD,
+        arr_size
+    );
+
+    if (ca->dimension == 1)
+        reset_rects(ca);
+}
 
 /**
  * 111 110 101 100 011 010 001 000
@@ -140,9 +160,7 @@ void rule_90(void *data) {
     new[0] = ca->cells[1];
     new[ca->len-1] = ca->cells[ca->len-2];
 
-    for (int i = 0; i < ca->len; i++) {
-        ca->rects[i].y += ca->rects[i].h;
-    }
+    next_layer(ca);
 
     ca->cells = new;
 }
@@ -171,9 +189,7 @@ void rule_110(void *data) {
     new[0] = ca->cells[1] | (ca->cells[0] ^ ca->cells[1]);
     new[ca->len-1] = ca->cells[ca->len-1];
 
-    for (int i = 0; i < ca->len; i++) {
-        ca->rects[i].y += ca->rects[i].h;
-    }
+    next_layer(ca);
 
     ca->cells = new;
 }
[…]

### 2018-09-21 simulation/src/main.c

@@ -130,7 +130,7 @@ int main(int argc, char *argv[]) {
 
     /* automata initialisation */
     int *config_1d = calloc(SIZE_1D, sizeof(int));
-    config_1d[400] = 1;
+    config_1d[SIZE_1D / 2] = 1;
 
     int *config = calloc(SIZE_2D*SIZE_2D, sizeof(int));
 
@@ -210,29 +210,30 @@ int main(int argc, char *argv[]) {
     int ticks = 0;
 
     /* main program execution */
-    active = automata[0];
+    // active = automata[0];
+    active = rule30;
 
     frame_start = SDL_GetTicks();
 
     while (running) {
 
+        // sprintf(gen_str, "%d", gens);
+        // if ((gen_len = strlen(gen_str)) > last_gen_len) {
+        //     last_gen_len = gen_len;
+        //     msg_rect.w += TXT_WIDTH;
+        //     msg_rect.x -= TXT_WIDTH;
+        // }
+
+        // surface = TTF_RenderText_Solid(font, gen_str, white);
+        // texture = SDL_CreateTextureFromSurface(renderer, surface);
+
+        /* do what it says on the tin */
         handle_input();
-
-        sprintf(gen_str, "%d", gens);
-        if ((gen_len = strlen(gen_str)) > last_gen_len) {
-            last_gen_len = gen_len;
-            msg_rect.w += TXT_WIDTH;
-            msg_rect.x -= TXT_WIDTH;
-        }
-
-        surface = TTF_RenderText_Solid(font, gen_str, white);
-        texture = SDL_CreateTextureFromSurface(renderer, surface);
 
         /* render automaton */
         render(active);
 
         if (!paused) {
-
             if (ticks > speed) {
 
                 /* update automaton */
@@ -246,7 +247,6 @@ int main(int argc, char *argv[]) {
                 ticks++;
 
             }
-
         } else {
             if (step) {
                 active->sim(active);
@@ -256,8 +256,8 @@ int main(int argc, char *argv[]) {
         }
 
         /* perhaps inefficient, but it works (TODO) */
-        SDL_DestroyTexture(texture);
-        SDL_FreeSurface(surface);
+        // SDL_DestroyTexture(texture);
+        // SDL_FreeSurface(surface);
 
         /* fix framerate at FRAME_RATE fps */
         frame_curr = SDL_GetTicks();
@@ -269,6 +269,7 @@ int main(int argc, char *argv[]) {
 
[…]
